---
title: "01b-General-Methylation-Patterns"
author: "Laura H Spencer"
date: "4/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries

```{r message=FALSE, warning=FALSE, results=FALSE}
list.of.packages <- c("tidyverse", "reshape2", "here", "scales", "dplyr", "corrplot") #add new libraries here 

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load all libraries 
lapply(list.of.packages, FUN = function(X) {
  do.call("require", list(X)) 
})
```

```{bash}
bedtools --version
```

## Get all CpG loci in genome 

```{r}
CpGs <- read.table(file = here::here("genome-features", "Olurida_v081_CG-motif.gff"))[,c(1,4,5)] %>% setNames(c("contig", "start", "end"))

# How many CpG loci are there in the genome? 
nrow(CpGs)
```

## Get methylation data 

Downloaded tab files with methylation proportions at each locus: 
- All loci b/f filtering: https://gannet.fish.washington.edu/seashell/bu-mox/scrubbed/020320-oly/zr1394_all_s456.CpG_report.merged_CpG_evidence.cov  
- 5x: https://gannet.fish.washington.edu/seashell/bu-mox/scrubbed/020320-oly/zr1394_all_s456_5x.tab  
- 10x: https://gannet.fish.washington.edu/seashell/bu-mox/scrubbed/020320-oly/zr1394_all_s456_10x.tab  
```{r}
allmeth <-read_delim(file=here::here("data", "zr1394_all_s456.CpG_report.merged_CpG_evidence.cov"), delim = "\t", col_names=FALSE) %>% 
  setNames(c("contig", "start", "end", "meth_perc", "methylated", "unmethylated"))
head(allmeth)

allmeth.5x <-read_delim(file=here::here("data", "zr1394_all_s456_5x.tab"), delim = "\t", col_names=FALSE) %>% 
  setNames(c("contig", "start", "end", "meth_perc", "methylated", "unmethylated"))
head(allmeth.5x)

# allmeth.10x <-read_delim(file=here::here("data", "zr1394_all_s456_10x.tab"), delim = "\t", col_names=FALSE) %>% 
#   setnames(c("contig", "start", "end", "meth_perc", "methylated", "unmethylated"))
# head(allmeth.10x)
```

### Call methylation status based on % methylation 

A locus is methylated if 50% or greater reads are methylated, that is they were unconverted after bisulfite treatment (in this dataset they are "C's") (Gavery and Roberts 2013; Olson and Roberts 2013). I will therefore determine methylation status using the numCs/coverage, where methylated = any greater or equal to 50%. 

```{r}
allmeth <- allmeth %>% 
  mutate(methyl.status=factor(ifelse(meth_perc >= 50, 'methylated', "unmethylated"))) %>%
  mutate(coverage=methylated+unmethylated)

allmeth.5x <- allmeth.5x %>% 
  mutate(methyl.status=factor(ifelse(meth_perc >= 50, 'methylated', "unmethylated"))) %>%
  mutate(coverage=methylated+unmethylated)

# allmeth.10x <- allmeth.10x %>% 
#   mutate(methyl.status=factor(ifelse(meth_perc >= 50, 'methylated', "unmethylated")))
```

## Calculate summary stats that Yaamini reports in her paper 
Work with 5x coverage file 

```{r}
# How many of loci with 5x coverage were methylated (>=50% reads methylated)? 
nrow(subset(allmeth, coverage>4 & meth_perc>=50))/nrow(subset(allmeth, coverage>4))*100

# How many of loci with 5x coverage were sparsely methylated?
nrow(subset(allmeth, coverage>4 & meth_perc<50 & meth_perc>=10))/nrow(subset(allmeth, coverage>4))*100

# How many of loci with 5x coverage were sparsely methylated?
nrow(subset(allmeth, coverage>4 & meth_perc<10))/nrow(subset(allmeth, coverage>4))*100

# How many loci are methylated with 5x coverage, compared to all CpG loci in genome? 
nrow(subset(allmeth, coverage>4 & meth_perc>=50))/nrow(CpGs)*100

nrow(allmeth)/nrow(CpGs)*100 #we have data for ~31% of CpGs in genome 
nrow(subset(allmeth, methyl.status=="methylated"))/nrow(CpGs)*100 #w/o filtering for coverage, 18% of CpGs in 

nrow(allmeth.5x)/nrow(CpGs)*100 #after filtering for 5x coverage, we have data for ~7.4% of CpGs in genome 
nrow(subset(allmeth.5x, methyl.status=="methylated"))/nrow(CpGs)*100 #after filtering for coverage, 6.7% of CpGs in genome are methylated

nrow(subset(allmeth.5x, methyl.status=="methylated"))/nrow(allmeth)*100
```

#### How many methylated & unmethylated sites are there before filtering for coverage? 

```{r}
# Mean proportion methylated 
mean(allmeth$meth_perc)

# Proportion of loci with 0% methylation 
sum(allmeth$meth_perc == 0)/nrow(allmeth)

# Frequency distribution of % methylation 
hist(allmeth$meth_perc, col="gray85", 
     xlab="% methylation per base", 
     main="% methylation, no coverage minimum (all samples)") 
```

#### How many methylated & unmethylated sites are there with 5x coverage? 

```{r}
# Mean proportion methylated 
mean(allmeth.5x$meth_perc)

# Proportion of loci with 0% methylation 
sum(allmeth.5x$meth_perc == 0)/nrow(allmeth.5x)

# Frequency distribution of % methylation 
hist(allmeth.5x$meth_perc, col="gray85", 
     xlab="% methylation per base", 
     main="% methylation, 5x minimum coverage (all samples)") 
```

<!-- #### How many methylated & unmethylated sites are there with 10x coverage?  -->

<!-- ```{r} -->
<!-- hist(allmeth.10x$meth_perc, col="gray85",  -->
<!--      xlab="% methylation per base",  -->
<!--      main="% methylation, 10x minimum coverage (all samples)")  -->
<!-- ``` -->

#### Write out bedfiles for methylated loci and unmethylated loci 

```{r}
# Extract methylated loci 
all_methylated_5x <- allmeth.5x %>% filter(methyl.status=="methylated")

# as .bed file 
write_delim(all_methylated_5x %>% dplyr::select(contig, start, end, meth_perc), here::here("analyses", "methylation-characteristics", "all_methylated_5x.bed"), delim = '\t', col_names = F)

# as R object 
save(all_methylated_5x, file="../analyses/methylation-characteristics/R-objects/all_methylated_5x")

# Extract unmethylated loci 
all_unmethylated_5x <- allmeth.5x %>% filter(methyl.status=="unmethylated")

# as .bed file 
write_delim(all_unmethylated_5x %>% dplyr::select(contig, start, end, meth_perc), here::here("analyses", "methylation-characteristics", "all_unmethylated_5.bed"), delim = '\t', col_names = F)

# as R object 
save(all_unmethylated_5x, file="../analyses/methylation-characteristics/R-objects/all_unmethylated_5x")
```

#### How many methylated loci are there? 

```{r}
nrow(all_methylated_5x)
```

#### How many unmethylated loci are there? 

```{r}
nrow(all_unmethylated_5x)
```

#### Identify locations of methylated loci 

```{bash}
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" >  ../analyses/methylation-characteristics/methylated-gene.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" >  ../analyses/methylation-characteristics/methylated-gene2kb.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.2kbflank-up.gff" >  ../analyses/methylation-characteristics/methylated-2kbflank-up.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.2kbflank-down.gff" >  ../analyses/methylation-characteristics/methylated-2kbflank-down.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" >  ../analyses/methylation-characteristics/methylated-exon.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" >  ../analyses/methylation-characteristics/methylated-CDS.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" >  ../analyses/methylation-characteristics/methylated-mRNA.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" >  ../analyses/methylation-characteristics/methylated-TE.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/methylation-characteristics/methylated-ASV.bed

bedtools intersect -v -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/methylation-characteristics/methylated-unknown.bed

#### Find the number of methylated loci that overlap with genes that have various # of isoforms 

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081_genes.1isoform.bed" >  ../analyses/methylation-characteristics/methylated-1isoform.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081_genes.2-5isoforms.bed" > ../analyses/methylation-characteristics/methylated-2-5isoforms.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081_genes.6-10isoforms.bed" > ../analyses/methylation-characteristics/methylated-6-10isoforms.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081_genes.11-20isoforms.bed" > ../analyses/methylation-characteristics/methylated-11-20isoforms.bed

bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081_genes.21-70isoforms.bed" > ../analyses/methylation-characteristics/methylated-21-70isoforms.bed
```

#### Start an R df to save feature counts 

First check to see if the file exists already. If it does, delete it, so I can re-create it from scratch 

```{r}
#Define the file name that will be deleted
fn <- "../analyses/methylation-characteristics/methylation-features.tab"
#Check its existence
if (file.exists(fn)) {
  #Delete file if it exists
  file.remove(fn)
}
```

Create a simple tab file with the # of methylated loci (i.e. lines) that are in each gene feature 

```{bash}
cat ../analyses/methylation-characteristics/methylated-mRNA.bed | wc -l  >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-CDS.bed | wc -l  >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-exon.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-gene.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-gene2kb.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-2kbflank-up.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-2kbflank-down.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-TE.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
grep exon ../analyses/methylation-characteristics/methylated-ASV.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-unknown.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/all_methylated_5x.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-1isoform.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-2-5isoforms.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-6-10isoforms.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-11-20isoforms.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
cat ../analyses/methylation-characteristics/methylated-21-70isoforms.bed | wc -l >> ../analyses/methylation-characteristics/methylation-features.tab
```

### Read in that tab file, add column with feature type 

```{r}
methdata.summary <- read_delim(here::here("analyses", "methylation-characteristics", "methylation-features.tab"), delim="\t", col_names=FALSE) %>%
  add_column(feature=c("mRNA", "CDS", "exon", "gene", "gene2kb", "2kbflank-up", "2kbflank-down","TE", "ASV", "unknown", "all", "1 isoform", "2-5 isoforms", "6-10 isoforms", "11-20 isoforms", "21-70 isoforms")) %>% 
  dplyr::rename(methylated=X1) %>% 
  mutate_at(vars(methylated), as.numeric)
```

#### Where are ALL loci located that we have data for? (this is methylated + unmethylated combined after filtering for 5x coverage) 

#### Use bedtools to find overlaps between all loci and gene features 

```{bash}
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.gene.gff" >  ../analyses/methylation-characteristics/all-gene.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" >  ../analyses/methylation-characteristics/all-gene2kb.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.2kbflank-up.gff" >  ../analyses/methylation-characteristics/all-2kbflank-up.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.2kbflank-down.gff" >  ../analyses/methylation-characteristics/all-2kbflank-down.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.exon.gff" >  ../analyses/methylation-characteristics/all-exon.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.CDS.gff" >  ../analyses/methylation-characteristics/all-CDS.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" >  ../analyses/methylation-characteristics/all-mRNA.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081_TE-Cg.gff" >  ../analyses/methylation-characteristics/all-TE.bed
bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/methylation-characteristics/all-ASV.bed
bedtools intersect -v -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/methylation-characteristics/all-unknown.bed

#### Find the number of all sequenced loci that overlap with genes that have various # of isoforms 

bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081_genes.1isoform.bed" >  "../analyses/methylation-characteristics/all-1isoform.bed"

bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081_genes.2-5isoforms.bed"  >  "../analyses/methylation-characteristics/all-2-5isoforms.bed"

bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081_genes.6-10isoforms.bed" >  "../analyses/methylation-characteristics/all-6-10isoforms.bed"

bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081_genes.11-20isoforms.bed" >  "../analyses/methylation-characteristics/all-11-20isoforms.bed"

bedtools intersect -wb -a "../data/zr1394_all_s456_5x.bedgraph" -b "../genome-features/Olurida_v081_genes.21-70isoforms.bed" >  "../analyses/methylation-characteristics/all-21-70isoforms.bed"
```

Generate a simple tab file with the # of loci that we have data for (5x coverage) that are in each gene feature 

First check to see if the file exists already. If it does, delete it, so I can re-create it from scratch 

```{r}
#Define the file name that will be deleted
fn <- "../analyses/methylation-characteristics/all5x-features.tab"
#Check its existence
if (file.exists(fn)) {
  #Delete file if it exists
  file.remove(fn)
}
```

Now, start building the .tab file with summary counts 

```{bash}
cat ../analyses/methylation-characteristics/all-mRNA.bed | wc -l  >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-CDS.bed | wc -l  >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-exon.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-gene.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-gene2kb.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-2kbflank-up.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-2kbflank-down.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-TE.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
grep exon ../analyses/methylation-characteristics/all-ASV.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-unknown.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../data/zr1394_all_s456_5x.bedgraph | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-1isoform.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-2-5isoforms.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-6-10isoforms.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-11-20isoforms.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
cat ../analyses/methylation-characteristics/all-21-70isoforms.bed | wc -l >> ../analyses/methylation-characteristics/all5x-features.tab
```

### Read in that tab file, add column with feature type 

```{r}
methdata.summary <- cbind(read_delim(here::here("analyses", "methylation-characteristics", "all5x-features.tab"), delim="\t", col_names=FALSE), methdata.summary) %>%
  rename(all5x=X1) %>% 
  mutate_at(vars(all5x), as.numeric)
```

### Where are all CpG loci in the Oly genome? 
This file is available for download here: https://owl.fish.washington.edu/halfshell/genomic-databank/Olurida_v081_CG-motif.gff 

```{bash}
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.gene.gff" > ../analyses/methylation-characteristics/CpGs-gene.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" >  ../analyses/methylation-characteristics/CpGs-gene2kb.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.2kbflank-up.gff" >  ../analyses/methylation-characteristics/CpGs-2kbflank-up.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.2kbflank-down.gff" >  ../analyses/methylation-characteristics/CpGs-2kbflank-down.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.exon.gff" > ../analyses/methylation-characteristics/CpGs-exon.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.CDS.gff" > ../analyses/methylation-characteristics/CpGs-CDS.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" > ../analyses/methylation-characteristics/CpGs-mRNA.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081_TE-Cg.gff" > ../analyses/methylation-characteristics/CpGs-TE.bed
bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" > ../analyses/methylation-characteristics/CpGs-ASV.bed
bedtools intersect -v -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" > ../analyses/methylation-characteristics/CpGs-unknown.bed


#### Find the number of CpG loci (candidate sites for methylation) that overlap with genes that have various # of isoforms 

bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081_genes.1isoform.bed" >  "../analyses/methylation-characteristics/CpGs-1isoform.bed"

bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081_genes.2-5isoforms.bed"  >  "../analyses/methylation-characteristics/CpGs-2-5isoforms.bed"

bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081_genes.6-10isoforms.bed" >  "../analyses/methylation-characteristics/CpGs-6-10isoforms.bed"

bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081_genes.11-20isoforms.bed" >  "../analyses/methylation-characteristics/CpGs-11-20isoforms.bed"

bedtools intersect -wb -a "../genome-features/Olurida_v081_CG-motif.gff" -b "../genome-features/Olurida_v081_genes.21-70isoforms.bed" >  "../analyses/methylation-characteristics/CpGs-21-70isoforms.bed"
```


### Add CpG counts to summary dataframe 

### Create a simple tab file with the # of loci (i.e. lines) that are in each gene feature 

First, delete if it already exists 

```{r}
#Define the file name that will be deleted
fn <- "../analyses/methylation-characteristics/CpGs-features.tab"
#Check its existence
if (file.exists(fn)) {
  #Delete file if it exists
  file.remove(fn)
}
```

```{bash}
cat ../analyses/methylation-characteristics/CpGs-mRNA.bed | wc -l  >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-CDS.bed | wc -l  >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-exon.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-gene.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-gene2kb.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-2kbflank-up.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-2kbflank-down.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-TE.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
grep exon ../analyses/methylation-characteristics/CpGs-ASV.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-unknown.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../genome-features/Olurida_v081_CG-motif.gff | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-1isoform.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-2-5isoforms.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-6-10isoforms.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-11-20isoforms.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
cat ../analyses/methylation-characteristics/CpGs-21-70isoforms.bed | wc -l >> ../analyses/methylation-characteristics/CpGs-features.tab
```

### Read in that tab file, add column with feature type 

```{r}
print(methdata.summary <- cbind(read_delim(here::here("analyses", "methylation-characteristics", "CpGs-features.tab"), delim="\t", col_names=FALSE), methdata.summary) %>%
  rename(CpGs=X1) %>% 
  mutate_at(vars(CpGs), as.numeric))
```

### Prep summary df that contains locations of methylated loci & all loci, and % of those loci that are located within known features 

```{r}
methdata.summary <- methdata.summary %>% 
  
  #add row with # loci that flank genes (up & down)
  rbind(c(methdata.summary[methdata.summary$feature=="gene2kb","CpGs"]-methdata.summary[methdata.summary$feature=="gene","CpGs"],
          methdata.summary[methdata.summary$feature=="gene2kb","all5x"]-methdata.summary[methdata.summary$feature=="gene","all5x"],
          methdata.summary[methdata.summary$feature=="gene2kb","methylated"]-methdata.summary[methdata.summary$feature=="gene","methylated"], 
          "geneflank2kb")) %>% 
  
  #add row with # loci in introns (# in genes minus # in exons)
  rbind(c(methdata.summary[methdata.summary$feature=="gene","CpGs"]-methdata.summary[methdata.summary$feature=="exon","CpGs"],
          methdata.summary[methdata.summary$feature=="gene","all5x"]-methdata.summary[methdata.summary$feature=="exon","all5x"],
          methdata.summary[methdata.summary$feature=="gene","methylated"]-methdata.summary[methdata.summary$feature=="exon","methylated"],
          "intron")) %>% 

  #convert all columns to numeric (except feature column)
  mutate_at(vars(-feature), funs(as.numeric)) %>% 
  
  #calculate % of loci that overlap with known features 
  mutate(CpGperc=CpGs/methdata.summary[methdata.summary$feature=="all", "CpGs"], 
         all5xperc=all5x/methdata.summary[methdata.summary$feature=="all", "all5x"], 
         methperc=methylated/methdata.summary[methdata.summary$feature=="all", "methylated"]) 

# Save summary object 
save(methdata.summary, file="../analyses/methylation-characteristics/R-objects/loci.summary")

methdata.summary.long <- cbind(melt(methdata.summary[,c("feature", "methylated", "all5x", "CpGs")], 
                                  variable.name = "analysis", value.name = "count"),
                             melt(methdata.summary[,c("feature", "methperc","all5xperc", "CpGperc")], 
                                  variable.name = "analysis", value.name = "percent"))[,c(1,2,3,6)] %>%
  mutate(feature=fct_relevel(as.factor(feature), c("gene", "gene2kb", "exon", "intron", "geneflank2kb", "2kbflank-up", "2kbflank-down", "CDS", "mRNA", "TE", "ASV", "1 isoform", "2-5 isoforms", "6-10 isoforms", "11-20 isoforms", "21-70 isoforms", "unknown", "all")),
         analysis=fct_relevel(as.factor(analysis), c("CpGs", "all5x", "methylated")))
```

## Conduct chi-squared test of homogeneity 

The null hypothesis is that loci distributions in the genome are the same between different categories.

### Is distribution of total CpGs vs. methylated CpGs the same? 

```{r}
meth.chisq <- methdata.summary %>% filter(feature=="gene" | feature=="exon" | feature=="intron" | feature=="2kbflank-up" | feature=="2kbflank-down" | feature=="TE" | feature=="unknown") %>% column_to_rownames("feature") %>% select(CpGs, methylated) %>% chisq.test() 

meth.chisq #no, they are different. P-value very low. 

meth.chisq$observed
meth.chisq$expected
corrplot(meth.chisq$residuals, is.cor = FALSE)
```

**Positive residuals are in blue.** Positive values in cells specify an attraction (positive association) between the corresponding row and column variables. 
--> Exons, genes, and introns are more likely to be methylated. 

**Negative residuals are in red.** This implies a repulsion (negative association) between the corresponding row and column variables. 
--> Transposable elements, and unknown regions are less likely. 

### Examine the contribution of each genome feature to the chi-square value (in percentage) 

```{r}
contrib <- 100*meth.chisq$residuals^2/meth.chisq$statistic
round(contrib, 3) 
corrplot(contrib, is.cor = FALSE)
```

The relative contribution of each cell to the total Chi-square score give some indication of the nature of the dependency between rows and columns of the contingency table. From the table/image above, it can be seen that the most contributing cells to the Chi-square are exon/methylated (41.62%), gene/methylated (21.39%), and unknown/methylated (26.7%). 

These cells contribute about 89.77% to the total Chi-square score and thus account for most of the difference between expected and observed values.


### Is distribution of total CpGs vs. methylated CpGs across genes with various # of isoforms the same? 

```{r}
meth.isoform.chisq <- methdata.summary %>% filter(feature=="1 isoform" | feature== "2-5 isoforms" | feature=="6-10 isoforms" | feature=="11-20 isoforms" | feature=="21-70 isoforms") %>% column_to_rownames("feature") %>% select(CpGs, methylated) %>% chisq.test() 

meth.isoform.chisq #no, they are different. P-value very low. 

meth.isoform.chisq$observed
meth.isoform.chisq$expected
corrplot(meth.isoform.chisq$residuals, is.cor = FALSE)
```

**Positive residuals are in blue.** Positive values in cells specify an attraction (positive association) between the corresponding row and column variables. 
--> genes with more isoforms (6+) are more likely to be methylated. 

**Negative residuals are in red.** This implies a repulsion (negative association) between the corresponding row and column variables. 
--> Genes with only 1 form are less likely to be methylated, compared to all CpG loci

### Examine the contribution of each genome feature to the chi-square value (in percentage) 

```{r}
contrib.isoform <- 100*meth.isoform.chisq$residuals^2/meth.isoform.chisq$statistic
round(contrib.isoform, 3) 
corrplot(contrib.isoform, is.cor = FALSE)
```
### Summary barplot showing where methylated loci and all CpG loci (potential methylated sites) overlap with known gene features (and unknown)

```{r}
ggplot(data=subset(methdata.summary.long, analysis!="all5x" & 
                     (feature=="gene" | feature=="exon" | feature=="intron" | feature=="2kbflank-up" | 
                        feature=="2kbflank-down" | feature=="TE" | feature=="unknown")), 
       aes(x=feature, y=percent, fill=analysis, label=percent(percent, accuracy = 0.1))) +  #prettyNum(count, big.mark = ",")
  geom_bar(stat="identity", position = "dodge", width = .5) +
scale_fill_manual(name = NULL, labels = c("All CpGs in genome", "Methylated CpGs"),
                  values=c("#a6cee3", "#1f78b4")) +
  ggtitle("% of all CpG loci and methylated loci that fall within genome features") +
  labs(y="% of Loci", x=NULL) +
  theme_minimal() + theme(legend.position ="bottom") + 
  geom_text(aes(label=percent(percent, accuracy = 0.1)), position=position_dodge(width=0.7), vjust=-0.1, size=3) +
  scale_x_discrete(
    labels=c(gene="Gene", exon="Exon", intron="Intron", `2kbflank-up`="5' flanking\nregion (-2kb)", 
             `2kbflank-down`="3' flanking\nregion (+2kb)", TE="Transposable\nelement",
             unknown="Unknown region\n of genome"))
```

### Summary barplot showing where methylated loci and all CpG loci (potential methylated sites) overlap with genes with various # of isoforms 

```{r}
ggplot(data=subset(methdata.summary.long, analysis!="all5x" & 
                     (feature=="1 isoform" | feature== "2-5 isoforms" | 
                        feature=="6-10 isoforms" | feature=="11-20 isoforms" | 
                        feature=="21-70 isoforms")), 
       aes(x=feature, y=percent, fill=analysis, label=percent(percent, accuracy = 0.1))) + 
  #prettyNum(count, big.mark = ",")
  geom_bar(stat="identity", position = "dodge", width = .5) +
scale_fill_manual(name = NULL, labels = c("All CpGs in genome", "Methylated CpGs"),
                  values=c("#a6cee3", "#1f78b4")) +
  ggtitle("% of all CpG loci and methylated loci that fall within genes with various # isoforms") +
  labs(y="% of Loci", x=NULL) +
    geom_text(aes(label=percent(percent, accuracy = 0.1)), position=position_dodge(width=0.7), vjust=-0.1, size=3) +
  theme_minimal() + theme(legend.position ="bottom") #+ 
```







# Boneyard 


### Summary STACKED barplot showing where methylated loci and all loci overlap with known gene features (and unknown)

```{r}
ggplot(data=subset(methdata.summary.long, feature=="2kbflank-up" | feature=="2kbflank-down" | feature=="exon" | feature=="intron" |feature=="TE" | feature=="unknown"), aes(x=analysis, y=percent, fill=feature, label=percent(percent, accuracy = 0.1))) +  #prettyNum(count, big.mark = ",")
  geom_bar(stat="identity", width = .5) +
 geom_bar(stat="identity", position="fill", width=0.5) + #use this instead for stacked bp to total 100%
scale_fill_manual(name = "Loci Location", labels = c("Exon", "Intron", "Upstream Flank (-2kb at 5')", "Downstream Flank (+2kb at 3')", "Transposable Elements", "Unknown Regions"),
                  values=c("#a6cee3", "#1f78b4", "#b2df8a","#33a02c", "#fb9a99", "gray")) +
  ggtitle("% of loci that overlap with genome features") +
  labs(y="% of Loci", x=NULL) +
  theme_minimal() + geom_text(size = 3, position = position_fill(vjust = 0.12)) + 
  scale_x_discrete(labels=c(CpGs="All CpG Loci",
    all5x="All Loci with\n5x Coverage", 
    "methylated" = "Methylated Loci"))
```



### DEFUNCT: Merge all .bam files into one, called "all.merged.bam", saved in the same directory (my external hard drive)

```
{bash, eval = FALSE}
samtools merge /Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/all.merged.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_1_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_2_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_3_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_4_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_5_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_6_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_7_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_8_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_9_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_10_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_11_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_12_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_13_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_14_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_15_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_16_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_17_s456_trimmed_bismark_bt2.deduplicated.sorted.bam \
/Volumes/Peach\ Backup/oly-mbdseq-bismark-files_sorted/zr1394_18_s456_trimmed_bismark_bt2.deduplicated.sorted.bam
```

### Download sorted alignment (bam) file that was generated from one large data file that included Oly MBD-Seq data from all samples in this study (concatenated prior to Bismark) - call this "super .bam"

### Create methylKit object the super .bam file 

```{r, eval = FALSE}
myobj_merged = processBismarkAln(location = "/Volumes/Bumblebee/paper-oly-mbdbs-gen/data/MBD-BAM-files/zr1394_all_s456_trimmed_bismark_bt2.deduplicated.sorted.bam", 
                                 assembly = "v081", read.context="CpG", mincov=2, sample.id="all_merged")
```

### Save methylRaw object to file

```{r, eval = FALSE}
save(myobj_merged, file = "../analyses/methylation-characteristics/R-objects/myobj_merged") 
```

### Read in methylRaw object, if neede d
```{r}
load("../analyses/methylation-characteristics/R-objects/myobj_merged") 
```

### Check out format of methylRaw object 

```{r}
head(myobj_merged)
```

```{r}
getMethylationStats(myobj_merged,plot=T,both.strands=TRUE)
```

```{r}
getCoverageStats(myobj_merged,plot=TRUE,both.strands=TRUE)
```

Create R dataframe object from methylRaw object 

```{r}
methdf_merged=getData(myobj_merged)
```

# Bypass MethylKit 

```{bash}
echo "No. of methylated loci located in genes:"
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" | wc -l 
```

```{bash}
echo "No. of methylated loci located in genes +/- 2kb flanks:"
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" | wc -l 
```

```{bash}
echo "No. of methylated loci located in exons:"
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" | wc -l 
```

```{bash}
echo "No. of methylated loci located in coding sequences:"
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" | wc -l 
```

```{bash}
echo "No. of methylated loci located in mRNA:"
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" | wc -l 
```

```{bash}
echo "No. of methylated loci located in transposable elements:"
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" | wc -l 
```

```{bash}
echo "No. of methylated loci located in alternative splice variants:"
bedtools intersect -wb -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l 
```

```{bash}
echo "No. of methylated loci not located in any known feature:"
bedtools intersect -v -a "../analyses/methylation-characteristics/all_methylated_5x.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l 
```

