---
title: "06-Analyzing MACAU results-rev"
author: "Laura H Spencer"
date: "9/12/2019"
output: html_document
---
## MACAU results - visualize and FDR correct  

Ran MACAU in a Jupyter Notebook. See [05-MACAU.ipynb](https://github.com/sr320/paper-oly-mbdbs-gen/blob/master/code/05-MACAU.ipynb)

In this notebook I assess patterns in methylated loci that MACAU identified as influence by oyster phenotype, aka oyster shell length. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load libraries 

```{r, message=FALSE, warning=FALSE, results=FALSE}
list.of.packages <- c("vegan", "dplyr", "factoextra", "FactoMineR", "tidyverse", "tibble", "reshape2", "gplots", "qvalue", "cluster") #add new libraries here 
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load all libraries 
lapply(list.of.packages, FUN = function(X) {
  do.call("require", list(X)) 
})
source("../resources/biostats.R")
```

#### Read in dataframe with MACAU assocation results 

```{r}
macau.20190812 <- read.table(file="../analyses/macau/output/20190812-macau.assoc.txt", sep = "\t", stringsAsFactors = F)
head(macau.20190812)
colnames(macau.20190812) <- macau.20190812[1,]
macau.20190812 <- macau.20190812[-1, ] 
macau.20190812$pvalue <- as.numeric(macau.20190812$pvalue)
hist(macau.20190812$pvalue)
tail(macau.20190812)
```

#### Correct for multiple comparisons 

From Mac's 2019 salmon paper, [doi:10.3390/genes10050356](10.3390/genes10050356)  
"Multiple hypothesis testing was performed on P-values extracted from the MACAU output for each CG site using the false discovery rate (FDR) approach used in the R package qvalue [45]. We considered a CG site to be differentially methylated (i.e., differentially methylated cytosine (DMC) if it passed a 10% FDR threshold, consistent with [46]."

```{r}
hist(macau.20190812$pvalue) # p-value distribution looks good (i.e. not u-shaped)
macau.qvalue <- qvalue(p = macau.20190812$pvalue, fdr.level = 0.1) # create qvalue object, setting FDR=10%
summary(macau.qvalue) 
100*(1- macau.qvalue$pi0) #% of all loci in df as truely sign.  
hist(macau.qvalue)

#if one wants to estimate the false discovery rate when calling all p-values ≤ 0.01 significant: 
max(macau.qvalue$qvalues[macau.qvalue$pvalues <= 0.01])

# The main purpose of the upper-left plot is to gauge the reliability of the π0 estimate, where the estimated π0 is plotted versus the tuning parameter λ. The remaining plots show how many tests are significant, as well as how many false positives to expect for each q-value cut-off.
plot(macau.qvalue)

# Given our FDR=10% level, how many are significant? 
summary(macau.qvalue$significant) #219 sign. (# TRUE)

# Add qvalue results to macau df 
macau.FDR <- cbind(macau.20190812, do.call(cbind.data.frame, macau.qvalue[c("qvalues", "pvalues", "significant")]))

#plot macau p-values against pvalues in the qvalue results - should all fall on 1:1 line (double check that pvalues are in same order) - 
plot(macau.FDR$pvalue, macau.FDR$pvalues) # looks good! 
```

#### Extract IDs for loci where MACAU indicates DML, FDR test (at 10%) deemed loci significantt 

```{r}
head(macau.sign <- subset(macau.FDR, significant=="TRUE"), n=20)
```

#### Read in count data 
Use dataframe that has the siteID, and original contig, start #, and end # info 

```{r}
head(meth.destranded.df <- read.table(file = "../analyses/meth.destranded.df.txt", sep = "\t", header=T, stringsAsFactors = F))
```

```{r}
#counts.tot.destrand <- read.table(file = "../analyses/counts-total-destrand.txt", header = T)
#counts.meth.destrand <- read.table(file = "../analyses/counts-meth-destrand.txt", header = T)
```

#### Merge MACAU + FDR results with raw count data 
**This constitutes a master dataframe** 

```{r}
# merge macau results with count dataframe, save to file 
head(macau.sign.counts <- merge(x=macau.sign, y=subset(meth.destranded.df, siteID %in% macau.sign$id), by.x="id", by.y="siteID"))
write.table(macau.sign.counts, file = "../analyses/combined-macau-FDR-counts.txt", sep = "\t", row.names = FALSE,  quote = FALSE) 
```

#### Pull out separate methylation count & coverage count dfs 

```{r}
macau.sign.tot.counts <- macau.sign.counts[, !(grepl("num", colnames(macau.sign.counts)))]
macau.sign.meth.counts <- macau.sign.counts[, !(grepl("numT|coverage", colnames(macau.sign.counts)))]
```

#### Create df where coverage counts <5 replaced with NA 

```{r}
macau.sign.tot.counts.5x <- macau.sign.tot.counts %>% 
  mutate_at(vars(contains('coverage')), funs(ifelse(. < 4, NA, .)))
```

#### Create % methylated dataframe with sign. MACAU loci 

```{r}
# duplicate macau methylated count dataframe to make % methylated df for 5x coverage (all <5x cov will be "NA"")
macau.sign.perc.meth.5x <- macau.sign.meth.counts

# divide macau methylated counts by total counts 
macau.sign.perc.meth.5x[,22:39] <- 100*(macau.sign.meth.counts[,22:39]/macau.sign.tot.counts.5x[,22:39])
head(macau.sign.perc.meth.5x[22:39])
rownames(macau.sign.perc.meth.5x) <- macau.sign.perc.meth.5x[,1]
```

#### Heatmaps of % methylated, with NA for instances that loci coverage <5x 

---> mbd samples #1-9 = Hood Canal population  
---> mbd samples #10-18 = South Sound population  

```{r}
par(cex.main=.8)
heatmap.2(as.matrix(macau.sign.perc.meth.5x[22:39]), main = "MACAU id'd % methylated\ngray=<5x cov", xlab = "mbd sample", ylab = "loci", na.color = "gray70", hclustfun = function(x) hclust(x, method="ward.D"), distfunc <- function(x) daisy(x,metric="gowers"), trace="none", dendrogram = "column" )
```

#### Prepare size data to plot in same order as this heat map 
This is laborious code with some room for error - needs fixing  

```{r}
size <- read.csv(file = "../data/mbd_size.csv", header=T, sep = "\t")
key <- read.csv("../data/sample-key.csv")
size.macau <- cbind(x=c(rep(1,times=18)), y=merge(y=size, x=key, by.y="Sample", by.x="SAMPLE"))
size.macau$pop <- as.factor(c(rep("HC", times=9), rep("SS",  times=9)))
size.macau$heat.order <- c(2,7,9,6,8,3,1,11,4,15,13,12,17,10,16,5,18,14)
mycols <- c("indianred", "darkcyan")
size.macau <- size.macau[order(size.macau$heat.order),]
```

#### Heatmaps of % methylation showing only loci with coverage <5x 
 
```{r}
par(cex.main=.8)

heatmap.2(as.matrix(macau.sign.perc.meth.5x[complete.cases(macau.sign.perc.meth.5x[22:39]), 22:39]), main = "MACAU id'd % methylation\nFDR=10%, loci cov=5x", xlab = "mbd sample", ylab = "loci", hclustfun = function(x) hclust(x, method="ward.D"), trace="none")

barplot(height = size.macau$y.Length,  names.arg = size.macau$y.MBD.FILENAME, col = mycols[size.macau$pop], cex.names =  0.8)
```

#### PCA with % methylated df, only with loci w/ 5x coverage across all samples 

```{r}
# Transpose methylated count df, remove any loci with NA values (i.e. <5x) 
macau.sign.perc.meth.5x.t <- macau.sign.perc.meth.5x[complete.cases(macau.sign.perc.meth.5x[22:39]), c(1,22:39)] %>% remove_rownames %>% gather(column, value, -id) %>% spread(id, value) %>% column_to_rownames(var="column")

# Run PCA on transposed matrix, scale=F creates variance-covariance matrix since all data is same units 
met.perc.pca <- prcomp(macau.sign.perc.meth.5x.t, scale=F) 
```

#### Inspect PCA axes, etc. 

```{r}
summary(met.perc.pca)
pca.eigenval(met.perc.pca) #The Proporation of Variance = %variance 
fviz_screeplot(met.perc.pca, addlabels = TRUE) #PC 1, 2 & 3 may be sign. 
ordi.monte(macau.sign.perc.meth.5x.t,ord="pca",dim=5) #only pc 1 sign. according to ordi.monte 
fviz_contrib(met.perc.pca, choice = "var", axes = 1, top=50)
fviz_contrib(met.perc.pca, choice = "var", axes = 2, top=50)
fviz_contrib(met.perc.pca, choice = "var", axes = 3, top=50)
```

#### Create PCA cluster plots (not with variables) 

```{r}
pops <- c("HC", rep("SS", times=9), rep("HC", times=8)) #create pop vector for color-coding 

fviz_pca_biplot(met.perc.pca, label="ind", invisible = "var", col.ind=pops, axes = c(1,2), pointsize=3) + ggtitle(label="PCA % methylated, loci id'd by MACAU, PC 1x2")

fviz_pca_biplot(met.perc.pca, label="ind",  invisible = "var",col.ind=pops, axes = c(1,3), pointsize=3) + ggtitle(label="PCA % methylated, loci id'd by MACAU, PC 1x3")

fviz_pca_biplot(met.perc.pca, label="ind", invisible = "var", col.ind=pops, axes = c(2,3), pointsize=3) + ggtitle(label="PCA % methylated, loci id'd by MACAU, PC 2x3")  #dimension 2 = pop separation 
```

