---
title: "06-Analyzing MACAU results-rev2"
author: "Laura H Spencer"
date: "1/29/2020"
output: html_document
---
## MACAU results - visualize and FDR correct  

Ran MACAU in a Jupyter Notebook. See [05-MACAU.ipynb](https://github.com/sr320/paper-oly-mbdbs-gen/blob/master/code/05-MACAU.ipynb)

In this notebook I assess patterns in methylated loci that MACAU identified as associated with the following, while also controlling for relatedness: 

  1) shell length, with shell whole wet weight as a covariate 
  2) population 

See my Jupyter notebook on these runs: [code/05-MACAU.ipynb](https://github.com/sr320/paper-oly-mbdbs-gen/blob/master/code/05-MACAU.ipynb)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries 

```{r, message=FALSE, warning=FALSE, results=FALSE}
list.of.packages <- c("vegan", "dplyr", "factoextra", "FactoMineR", "tidyverse", "tibble", "reshape2", "gplots", "qvalue", "cluster", "here", "gridExtra", "RColorBrewer") #add new libraries here 
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load all libraries 
lapply(list.of.packages, FUN = function(X) {
  do.call("require", list(X)) 
})
```

## Read in count data 
Use dataframes that have the siteID, and original contig, start #, and end # info 

```{r}
head(counts.filtered.meth <- read.table(file = here::here("analyses", "counts-filtered-meth.txt"), sep = "\t", header=T, stringsAsFactors = F))
head(counts.filtered.tot <- read.table(file = here::here("analyses", "counts-filtered-total.txt"), sep = "\t", header=T, stringsAsFactors = F))
```

# 1. Analyse MACAU results to assess loci where methylation is associated with shell length    
  - MACAU results which analyzed count data for loci where 8 of the 9 of samples in each population had **10x coverage** or greater (using `unite()` function in methylkit)  
  - MACAU was run with shell length as the predictor variable and whole wet weight was the covariate  

### Read in dataframe with MACAU assocation results 

```{r}
MACAU.results.length <- read.table(file=here::here("analyses","macau","output","20200205-macau.assoc.txt"), sep = "\t", stringsAsFactors = F)
colnames(MACAU.results.length) <- MACAU.results.length[1,]
MACAU.results.length <- MACAU.results.length[-1, ] 
MACAU.results.length$pvalue <- as.numeric(MACAU.results.length$pvalue)
hist(MACAU.results.length$pvalue)

length(MACAU.results.length$id)
length(counts.filtered.tot$siteID)

summary(MACAU.results.length$id == counts.filtered.tot$siteID) #confirm that macau results and original destranded dataframe are in same order 
MACAU.results.length.c <- cbind(counts.filtered.tot[,1],MACAU.results.length) %>%
   separate(`counts.filtered.tot[, 1]`, into = c("contig", "locus"), sep = "_")
write.table(MACAU.results.length.c, file = here::here("analyses", "macau", "MACAU.results.length.c.tab"), sep = "\t", row.names = FALSE,  quote = FALSE)
```

### Correct for multiple comparisons 

From Mac's 2019 salmon paper, [doi:10.3390/genes10050356](10.3390/genes10050356)  
"Multiple hypothesis testing was performed on P-values extracted from the MACAU output for each CG site using the false discovery rate (FDR) approach used in the R package qvalue [45]. We considered a CG site to be differentially methylated (i.e., differentially methylated cytosine (DMC) if it passed a 10% FDR threshold, consistent with [46]."

```{r}
hist(MACAU.results.length.c$pvalue) # p-value distribution looks good (i.e. not u-shaped)
MACAU.results.length <- qvalue(p = MACAU.results.length.c$pvalue, fdr.level = 0.1) # create qvalue object, setting FDR=10%
summary(MACAU.results.length) 
100*(1- MACAU.results.length$pi0) #% of all loci in df as truely sign.  
hist(MACAU.results.length)

#if one wants to estimate the false discovery rate when calling all p-values ≤ 0.01 significant: 
max(MACAU.results.length$qvalues[MACAU.results.length$pvalues <= 0.01])

# The main purpose of the upper-left plot is to gauge the reliability of the π0 estimate, where the estimated π0 is plotted versus the tuning parameter λ. The remaining plots show how many tests are significant, as well as how many false positives to expect for each q-value cut-off.
plot(MACAU.results.length)

# Given our FDR=10% level, how many are significant? 
summary(MACAU.results.length$significant) #15 sign. (# TRUE)

# Add qvalue results to macau df 
macau.FDR.length <- cbind(MACAU.results.length.c, do.call(cbind.data.frame, MACAU.results.length[c("qvalues", "pvalues", "significant")]))

#plot macau p-values against pvalues in the qvalue results - should all fall on 1:1 line (double check that pvalues are in same order) 
plot(macau.FDR.length$pvalue, macau.FDR.length$pvalues) # looks good! 
```

### Extract IDs for loci where MACAU indicates DML, FDR test (at 10%) deemed loci significantt 

```{r}
head(macau.sign.length <- subset(macau.FDR.length, significant=="TRUE"))
nrow(macau.sign.length) #number of significant loci = 4
```

### Merge MACAU + FDR results with raw count data 

Merge macau results with count dataframe, save to file.
**These constitute master dataframes** 

```{r}
macau.sign.length.meth.counts <- merge(x=macau.sign.length, y=subset(counts.filtered.meth, siteID %in% macau.sign.length$id), by.x="id", by.y="siteID")
macau.sign.length.tot.counts <- merge(x=macau.sign.length, y=subset(counts.filtered.tot, siteID %in% macau.sign.length$id), by.x="id", by.y="siteID")
write.table(macau.sign.length.meth.counts, file = here::here("analyses", "combined-macau-FDR-meth-counts.length.txt"), sep = "\t", row.names = FALSE,  quote = FALSE) 
write.table(macau.sign.length.tot.counts, file = here::here("analyses", "combined-macau-FDR-tot-counts.length.txt"), sep = "\t", row.names = FALSE,  quote = FALSE) 
```

### Create df where coverage counts <5 replaced with NA, & create % methylated dataframe with sign. MACAU loci 

```{r}
macau.sign.length.tot.counts.na <- macau.sign.length.tot.counts %>% 
  mutate_at(vars(contains('coverage')), funs(ifelse(. < 5, NA, .)))
macau.sign.length.perc.meth <- macau.sign.length.meth.counts
macau.sign.length.perc.meth[,20:37] <- 100*(macau.sign.length.meth.counts[,20:37]/macau.sign.length.tot.counts.na[,20:37])
```

### Heatmaps  

#### Prep order vectors

```{r}
size <- read.csv(file = here::here("data","mbd_size.csv"), header=T, sep = "\t")
key <- read.csv(file=here::here("data","sample-key.csv"))
size.macau <- cbind(x=c(rep(1,times=18)), y=merge(y=size, x=key, by.y="Sample", by.x="SAMPLE")) %>%
  mutate(population=as.factor(c(rep("HC", times=9), rep("SS",  times=9)))) %>%
  mutate(sample=paste("numCs", y.MBD.FILENAME, sep="")) %>%
  arrange(y.Length)
mycols <- c("firebrick3", "dodgerblue3")

tree.order <- paste("numCs",c(15,17,12,10,11,13,18,14,16,8,1,6,9,2,4,5,3,7), sep="")
size.order <- paste("numCs",c(size.macau[order(size.macau$y.Length),"y.MBD.FILENAME"]), sep="")

pop.colors.tree <- c(rep("dodgerblue3", times=9), rep("firebrick3", times=9))
pop.colors.length <- c("dodgerblue3", rep("firebrick3", times=7), rep("dodgerblue3", times=3), "firebrick3","dodgerblue3","firebrick3",rep("dodgerblue3", times=4))

macau.sign.length.perc.meth <- macau.sign.length.perc.meth %>%
  mutate(mean.HC = rowMeans(macau.sign.length.perc.meth[,20:28])) %>%
  mutate(mean.SS = rowMeans(macau.sign.length.perc.meth[,29:37])) %>%
  mutate(HC.SS.ratio = mean.HC/mean.SS) %>%
  arrange(HC.SS.ratio)
rownames(macau.sign.length.perc.meth) <- macau.sign.length.perc.meth[,1]
```

#### Heatmap of loci associated with shell length, ordered by the heatmap cluster analysis 

```{r}
par(cex.main=.8)
heatmap.2(as.matrix(macau.sign.length.perc.meth[, grep(c("numCs"), colnames(macau.sign.length.perc.meth))]), 
          main = "MACAU id'd % methylated\n5x cov\nOrdered by cluster analysis", 
          xlab = NA, ylab = NA, 
          hclustfun = function(x) hclust(x, method="ward.D"), distfunc <- function(x) daisy(x,metric="correlation"), trace="none", dendrogram = "column", na.color = "white",
          col = colorRampPalette(brewer.pal(4,"YlGnBu")), colCol = c(rep("firebrick3", times=9), rep("dodgerblue3", times=9)), cexRow = 0.6)
```

#### Heatmap of loci associated with shell length, ordered by cluster analysis from MethylKit 

```{r}
par(cex.main=.8)
heatmap.2(as.matrix(macau.sign.length.perc.meth[, grep(c("numCs"), colnames(macau.sign.length.perc.meth))][tree.order]), 
          main = "MACAU id'd % methylated\ngray=<5x cov\nOrdered by MethylKit tree", 
         xlab = NA, 
         na.color = "white",trace="none", dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(brewer.pal(4,"YlGnBu")), colCol = pop.colors.tree, cexRow = 0.6)
```

#### Heatmap of loci associated with shell length, ordered by shell length 

```{r}
par(cex.main=.8)
heatmap.2(as.matrix(macau.sign.length.perc.meth[, grep(c("numCs"), colnames(macau.sign.length.perc.meth))][size.order]), 
          main = "MACAU id'd % methylation\nFDR=10%\nOrdered by Length", 
          xlab = NA, ylab = NA, 
          trace="none", dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(brewer.pal(4,"YlGnBu")), colCol = pop.colors.length, cexRow = 0.6)
```

#### Barplot of shell length, ordered by length 

```{r}
par(oma=c(0,0,5,0))
barplot(size.macau[order(size.macau$y.Length),]$y.Length,  names.arg = size.macau[order(size.macau$y.Length),]$y.MBD.FILENAME, col = mycols[size.macau[order(size.macau$y.Length),]$pop], cex.names =  0.8, main = "Shell Length (mm)", las=2, space=1) #could add this to cut yaxis off at 10mm: (ylim=c(10,40), xpd = FALSE)
abline(h = 0, col = "black", lwd = 1)
```

### Create BED-type files

```{r}
macau.sign.length.perc.meth %>% 
  mutate(end = locus) %>% 
  rename(start = locus) %>%
  dplyr::select(c("contig", "start", "end")) %>%
write_delim(here::here("analyses","macau", "macau.sign.length.perc.meth.bed"),  delim = '\t', col_names = FALSE)

# All loci fed into/from macau 
MACAU.results.length.c %>%
  mutate(end = locus) %>% 
  rename(start = locus) %>%
  dplyr::select(c("contig", "start", "end")) %>%
write_delim(here::here("analyses","macau", "macau-all-loci.bed"),  delim = '\t', col_names = FALSE)

# All loci fed into/from macau with q-values  
macau.FDR.length %>%
write_delim(here::here("analyses","macau", "macau-all-loci_qvalues.tab"),  delim = '\t', col_names = FALSE)
```


# Analyze MACAU results to assess loci where methylation is associated with population  

  - MACAU results which analyzed count data for loci where 8 of the 9 of samples in each population had **10x coverage** or greater. 
  - MACAU was run with binary predictor variables that were 1=Hood Canal and 2=South Sound 

```{r}
MACAU.results.pop <- read.table(file=here::here("analyses","macau","output","20200310-macau.assoc.txt"), sep = "\t", stringsAsFactors = F)
colnames(MACAU.results.pop) <- MACAU.results.pop[1,]
MACAU.results.pop <- MACAU.results.pop[-1, ] 
MACAU.results.pop$pvalue <- as.numeric(MACAU.results.pop$pvalue)
hist(MACAU.results.pop$pvalue)

# Check to see if contig IDs are in the same order for macau results and count dataframe 
length(MACAU.results.pop$id) - length(counts.filtered.tot$siteID) # not the same, 63 fewer in MACAU results 
subset(counts.filtered.tot, !(siteID %in% c(MACAU.results.pop$id)))

#identify code to see whether contig IDs are in the same order for macau results and count dataframe  
summary(subset(counts.filtered.tot, (siteID %in% c(MACAU.results.pop$id)))[,"siteID"] == MACAU.results.pop$id)

# apply above code 
MACAU.results.pop.c <- cbind(
  subset(counts.filtered.tot, (siteID %in% c(MACAU.results.pop$id)))[,1],
  MACAU.results.pop) %>%
   separate(`subset(counts.filtered.tot, (siteID %in% c(MACAU.results.pop$id)))[, `, 
            into = c("contig", "locus"), sep = "_")

write.table(MACAU.results.pop.c, file = here::here("analyses", "macau", "MACAU.results.pop.c.tab"), sep = "\t", row.names = FALSE,  quote = FALSE)
```

### Correct for multiple comparisons 

```{r}
hist(MACAU.results.pop.c$pvalue) # p-value distribution looks good (i.e. not u-shaped)
macau.qvalue <- qvalue(p = MACAU.results.pop.c$pvalue, fdr.level = 0.1) # create qvalue object, setting FDR=10%
summary(macau.qvalue) 
100*(1- macau.qvalue$pi0) #% of all loci in df as truely sign.  
hist(macau.qvalue)
max(macau.qvalue$qvalues[macau.qvalue$pvalues <= 0.01])
plot(macau.qvalue)
summary(macau.qvalue$significant) #0 sign. (# TRUE)
```

### Stop analysis, there are no significant loci associated with population, after controlling for relatedness. 