---
title: "06-Annotating DMLs and MACAU loci"
author: "Laura H Spencer"
date: "11/20/2019"
output: html_document
---

### Load libraries 

```{r setup, message=FALSE, warning=FALSE, results=FALSE}
list.of.packages <- c("tidyverse", "reshape2", "here", "scales") #add new libraries here 
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load all libraries 
lapply(list.of.packages, FUN = function(X) {
  do.call("require", list(X)) 
})
sessionInfo()
```

## Use bedtools to see where DMLs and MACAU loci are located.

DMLs between the Olympia oyster populations, Hood Canal and South Sound, were identified using MethylKit. File is: [analyses/dml25.bed](https://github.com/sr320/paper-oly-mbdbs-gen/blob/master/analyses/dml25.bed) 

MACAU was used to identify loci at which methylation is associated with a phenotype, in our case shell length, while controlling for relatedness. 
### Locations of feature files 

[Olurida_v081-20190709.gene.gff](https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/genome-features/Olurida_v081-20190709.gene.gff) - genes, gene = "../genome-features/Olurida_v081-20190709.gene.gff"   
[Olurida_v081-20190709.gene.2kbslop.gff](https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/genome-features/Olurida_v081-20190709.gene.2kbslop.gff) - genes +/- 2kb, gene2kb = "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff"       
[Olurida_v081-20190709.CDS.gff](https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/genome-features/Olurida_v081-20190709.CDS.gff) - Coding regions of genes, CDS = "../genome-features/Olurida_v081-20190709.CDS.gff"      
[Olurida_v081-20190709.exon.gff](https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/genome-features/Olurida_v081-20190709.exon.gff) - Exons, exon = "../genome-features/Olurida_v081-20190709.exon.gff"     
[Olurida_v081-20190709.mRNA.gff](https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/genome-features/Olurida_v081-20190709.mRNA.gff) - mRNA, mRNA = "../genome-features/Olurida_v081-20190709.mRNA.gff"      
[Olurida_v081_TE-Cg.gff](https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/genome-features/Olurida_v081_TE-Cg.gff) - Transposable elements, TE = "../genome-features/Olurida_v081_TE-Cg.gff"    
[20190709-Olurida_v081.stringtie.gtf](https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/genome-features/20190709-Olurida_v081.stringtie.gtf) - alternative splice variants, ASV = "../genome-features/20190709-Olurida_v081.stringtie.gtf"   

### Background files used for MACAU and DMLs 

AllLociMACAU = "../analyses/macau/macau-all-loci.bed"
AllLociDMLs = "../analyses/DMLs/mydiff-all.bed"

### File with MACAU loci 
macau = "../analyses/macau/macau.sign.length.perc.meth.bed"

### File with DML loci 

DML = "../analyses/DMLs/dml25.bed"

### Identify MACAU and DML loci in each genome feature using Bedtools intersect 

I will use `bedtools` to identify where DML and MACAU loci intersect with known genome features. 

`bedtool intersect` options to use:  
`-u` - Write the original A entry _once_ if _any_ overlaps found in B, _i.e._ just report the fact >=1 hit was found  
`-a` - File A  
`-b` - File B  

## 1. DMLs 

```{bash}
bedtools intersect -wb -a "../analyses/DMLs/dml25.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" >  ../analyses/DMLs/DML-gene.bed
bedtools intersect -wb -a "../analyses/DMLs/dml25.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" >  ../analyses/DMLs/DML-gene2kb.bed
bedtools intersect -wb -a "../analyses/DMLs/dml25.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" >  ../analyses/DMLs/DML-exon.bed
bedtools intersect -wb -a "../analyses/DMLs/dml25.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" >  ../analyses/DMLs/DML-CDS.bed
bedtools intersect -wb -a "../analyses/DMLs/dml25.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" >  ../analyses/DMLs/DML-mRNA.bed
bedtools intersect -wb -a "../analyses/DMLs/dml25.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" >  ../analyses/DMLs/DML-TE.bed
bedtools intersect -wb -a "../analyses/DMLs/dml25.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/DMLs/DML-ASV.bed
bedtools intersect -v -a "../analyses/DMLs/dml25.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/DMLs/DML-unknown.bed
```

### Background loci features that were used to identify DMLs  

```{bash}
bedtools intersect -wb -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" >  ../analyses/DMLs/AllLociDMLs-gene.bed
bedtools intersect -wb -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" >  ../analyses/DMLs/AllLociDMLs-gene2kb.bed
bedtools intersect -wb -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" >  ../analyses/DMLs/AllLociDMLs-exon.bed
bedtools intersect -wb -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" >  ../analyses/DMLs/AllLociDMLs-CDS.bed
bedtools intersect -wb -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" >  ../analyses/DMLs/AllLociDMLs-mRNA.bed
bedtools intersect -wb -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" >  ../analyses/DMLs/AllLociDMLs-TE.bed
bedtools intersect -wb -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/DMLs/AllLociDMLs-ASV.bed
bedtools intersect -v -a "../analyses/DMLs/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/DMLs/AllLociDMLs-unknown.bed
```

## 2. MACAU Loci 

```{bash}
bedtools intersect -wb -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" >  ../analyses/macau/macau-gene.bed
bedtools intersect -wb -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" >  ../analyses/macau/macau-gene2kb.bed
bedtools intersect -wb -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" >  ../analyses/macau/macau-exon.bed
bedtools intersect -wb -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" >  ../analyses/macau/macau-CDS.bed
bedtools intersect -wb -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" >  ../analyses/macau/macau-mRNA.bed
bedtools intersect -wb -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" >  ../analyses/macau/macau-TE.bed
bedtools intersect -wb -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/macau/macau-ASV.bed
bedtools intersect -v -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/macau/macau-unknown.bed
```

###  Background loci used for MACAU  

```{bash}
bedtools intersect -wb -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" >  ../analyses/macau/AllLociMACAU-gene.bed
bedtools intersect -wb -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" >  ../analyses/macau/AllLociMACAU-gene2kb.bed
bedtools intersect -wb -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" >  ../analyses/macau/AllLociMACAU-exon.bed
bedtools intersect -wb -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" >  ../analyses/macau/AllLociMACAU-CDS.bed
bedtools intersect -wb -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" >  ../analyses/macau/AllLociMACAU-mRNA.bed
bedtools intersect -wb -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" >  ../analyses/macau/AllLociMACAU-TE.bed
bedtools intersect -wb -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/macau/AllLociMACAU-ASV.bed
bedtools intersect -v -a "../analyses/macau/macau-all-loci.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" >  ../analyses/macau/AllLociMACAU-unknown.bed
```

## Prepare blastx annotation files to merge with DML and MACAU results 

#### Download oly blastx file
NOTE: remove "eval=FALSE" to execute this code chunk 

```{bash, eval = FALSE}
#curl https://raw.githubusercontent.com/sr320/paper-oly-mbdbs-gen/master/analyses/Olgene_blastx_uniprot.05.tab > ../data/Olgene_blastx_uniprot.05.tab
```

#### convert pipes to tab

```{bash}
tr '|' '\t' < ../data/Olgene_blastx_uniprot.05.tab \
> ../data/Olgene_blastx_uniprot.05-20191122.tab
wc -l ../data/Olgene_blastx_uniprot.05-20191122.tab
```

#### Reduce the number of columns using awk. Sort, and save as a new file.

```{bash}
awk -v OFS='\t' '{print $1, $3, $13}' \
< ../data/Olgene_blastx_uniprot.05-20191122.tab | sort \
> ../data/Olgene_blastx_uniprot.05-20191122-sort.tab
wc -l ../data/Olgene_blastx_uniprot.05-20191122-sort.tab
```

#### Preview blastx annotation files file that will be joined with feature lists to annotate 

```{bash}
head ../data/Olgene_blastx_uniprot.05-20191122-sort.tab
```

#### Read blastx annotation file to object 

```{r}
uniprot <- read_delim(here::here("data", "Olgene_blastx_uniprot.05-20191122-sort.tab"), delim="\t", col_names = FALSE) %>%
  setNames(c("gene", "UniprotID", "unknown")) %>% 
  separate(gene, into = c("contig.feat", "start.feat", "end.feat"), sep = '[:-]') %>%
  mutate_at(c("start.feat", "end.feat"),as.numeric)
```

# Summarizing Annotations 

## 1. DMLs - where are differentially methylated loci located? 

```{r, message=FALSE, warning=FALSE}
DMLfiles <- c("DML-CDS.bed", "DML-exon.bed", "DML-gene.bed", "DML-gene2kb.bed", "DML-mRNA.bed", "DML-TE.bed", "DML-ASV.bed", "DML-unknown.bed")
DML.features <- list()
for (i in c(1:8)) {
  DML.features[[i]] <- read_delim(here::here("analyses", "DMLs", DMLfiles[i]), delim = '\t', col_names = FALSE) %>% as_tibble()}
for (i in 1:7) {
  DML.features[[i]] <- DML.features[[i]] %>%
    setNames(c("contig.dml","start.dml","end.dml","score.dml","contig.feat", "source.feat","feature","start.feat","end.feat","unknown1","strand","unknown2","attribute")) %>%
mutate(ID=str_extract(attribute, "ID=(.*?);"),
       Parent=str_extract(attribute, "Parent=(.*?);"),
       Name=str_extract(attribute, "Name=(.*?);"),
       Alias=str_extract(attribute, "Alias=(.*?);"),
       AED=str_extract(attribute, "AED=(.*?);"),
       eAED=str_extract(attribute, "eAED=(.*?);"),
       Note=str_extract(attribute, "Note=(.*?);"),
       Ontology_term=str_extract(attribute, "Ontology_term=(.*?);"),
       Dbxref=str_extract(attribute, "Dbxref=(.*?);")
       ) %>%
mutate_at("feature", as.factor)
}
names(DML.features) <- c("DML.CDS", "DML.exon", "DML.gene","DML.gene2kb", "DML.mRNA", "DML.TE", "DML.ASV", "DML.unknown")
DML.features[["DML.unknown"]] <- 
  DML.features[["DML.unknown"]] %>% 
  setNames(c("contig.DML", "start.dml", "end.dml", "score.dml"))

DML.features[["DML.CDS"]] <- DML.features[["DML.CDS"]] %>% mutate_at("unknown2", as.character)
DML.features[["DML.TE"]] <- DML.features[["DML.TE"]] %>% mutate_at("unknown1", as.character)
DML.features[["DML.ASV"]] <- DML.features[["DML.ASV"]] %>% mutate_at("unknown1", as.character)
DML.features[["DML.gene2kb"]]$feature <- "gene2kb"
DML.features[["DML.unknown"]]$feature <- "unknown"
DML.features.df <- bind_rows(DML.features[-7])
print(DML.summary <- table(DML.features.df[c("feature")])) #Note: "similarity" refers to transposable elements; also NO alternative splice variants are included. 

# 
DML.features[[3]] %>% left_join(uniprot, by=c("contig.feat","start.feat", "end.feat")) 

#save DML loci feature df object to file, to use in notebook #12 
save(DML.features.df, file="../analyses/DMLs/R-objects/DML.features.df")
save(DML.summary, file="../analyses/DMLs/R-objects/DML.summary")
```

## 2. MACAU - where are size-associated methylated loci located? 

```{r, message=FALSE, warning=FALSE}
macau.files <- c("macau-CDS.bed", "macau-exon.bed", "macau-gene.bed", "macau-gene2kb.bed", "macau-mRNA.bed", "macau-ASV.bed", "macau-TE.bed", "macau-unknown.bed")
macau.features <- list()
for (i in c(1:8)) {
  macau.features[[i]] <- read_delim(here::here("analyses", "macau", macau.files[i]), delim = '\t', col_names = FALSE) %>% as_tibble()}
for (i in 1:6) {
  macau.features[[i]] <- macau.features[[i]] %>%
    setNames(c("contig.macau","start.macau","end.macau","contig.feat", "source.feat","feature","start.feat","end.feat","unknown1","strand","unknown2","attribute")) %>%
mutate(ID=str_extract(attribute, "ID=(.*?);"),
       Parent=str_extract(attribute, "Parent=(.*?);"),
       Name=str_extract(attribute, "Name=(.*?);"),
       Alias=str_extract(attribute, "Alias=(.*?);"),
       AED=str_extract(attribute, "AED=(.*?);"),
       eAED=str_extract(attribute, "eAED=(.*?);"),
       Note=str_extract(attribute, "Note=(.*?);"),
       Ontology_term=str_extract(attribute, "Ontology_term=(.*?);"),
       Dbxref=str_extract(attribute, "Dbxref=(.*?);")
       ) %>%
mutate_at("feature", as.factor)
}
names(macau.features) <- c("macau.CDS", "macau.exon","macau.gene", "macau.gene2kb","macau.mRNA", "macau.ASV", "macau.TE", "macau.unknown")
macau.features[["macau.CDS"]] <- macau.features[["macau.CDS"]] %>% mutate_at("unknown2", as.character)
macau.features[["macau.ASV"]] <- macau.features[["macau.ASV"]] %>% mutate_at("unknown1", as.character)
macau.features[["macau.gene2kb"]]$feature <- "gene2kb"
#macau.features[["macau.unknown"]]$feature <- "unknown"
macau.features.df <- bind_rows(macau.features[-6:-7])
print(macau.summary <- table(macau.features.df[c("feature")])) #NOTE this does not contain alternative splice variants. There are zero transposable elements and unknown loci.  

save(macau.features.df, file="../analyses/macau/R-objects/macau.features.df")
save(macau.summary, file="../analyses/macau/R-objects/macau.summary")

#macau.features[["macau.gene"]] %>% inner_join(uniprot, by=c("contig.feat","start.feat", "end.feat"))  #not sure if this is working?
```
### Barplots of mean % methylation by population, MACAU loci in known features  

```{r}
load(here::here("analyses", "methylation-filtered", "R-objects", "meth_filter_reshaped"))

meth_filter_calcs <- meth_filter_reshaped %>% 
  group_by(population, chr, start) %>% 
  dplyr::summarise(
    mean_percMeth = mean(percMeth, na.rm=TRUE),
    sd_percMeth=sd(percMeth, na.rm=TRUE),
    n()) 

meth_filter_calcs %>% 
      filter(chr %in% macau.features[["macau.gene2kb"]]$contig.macau &
           start %in% macau.features[["macau.gene2kb"]]$start.macau) %>%
  ggplot(aes(x = population, y = mean_percMeth, fill = population, 
                         label=paste0(round(mean_percMeth, digits = 2), "%"))) + 
      geom_bar(stat="identity", width = 0.5) + ylim(0,110) +
      geom_pointrange(aes(ymin=mean_percMeth, 
                        ymax=mean_percMeth+sd_percMeth, width=0.15)) + 
      geom_text(size=3, vjust=-0.5, hjust=1.25) +
      theme_light() + ggtitle("") + facet_wrap(~chr) + 
    scale_fill_manual(values=c("firebrick3","dodgerblue3"))
```

## 3. All background loci used for DMLs & MACAU 

### Where are DML background loci located? 

```{r, message=FALSE, warning=FALSE}
allLociDMLfiles <- c("AllLociDMLs-CDS.bed", "AllLociDMLs-exon.bed", "AllLociDMLs-gene.bed",  "AllLociDMLs-gene2kb.bed", "AllLociDMLs-mRNA.bed", "AllLociDMLs-TE.bed", "AllLociDMLs-ASV.bed", "AllLociDMLs-unknown.bed")
allLociDML.features <- list()
for (i in c(1:8)) {
  allLociDML.features[[i]] <- read_delim(here::here("analyses","DMLs", allLociDMLfiles[i]), delim = '\t', col_names = FALSE) %>% as_tibble()}
for (i in 1:7) {
  allLociDML.features[[i]] <- allLociDML.features[[i]] %>%
    setNames(c("contig.allLoci","start.allLoci","end.allLoci", "unknown", "contig.feat", "source.feat","feature","start.feat","end.feat","unknown1","strand","unknown2","attribute")) %>%
mutate(ID=str_extract(attribute, "ID=(.*?);"),
       Parent=str_extract(attribute, "Parent=(.*?);"),
       Name=str_extract(attribute, "Name=(.*?);"),
       Alias=str_extract(attribute, "Alias=(.*?);"),
       AED=str_extract(attribute, "AED=(.*?);"),
       eAED=str_extract(attribute, "eAED=(.*?);"),
       Note=str_extract(attribute, "Note=(.*?);"),
       Ontology_term=str_extract(attribute, "Ontology_term=(.*?);"),
       Dbxref=str_extract(attribute, "Dbxref=(.*?);")
       ) %>%
mutate_at("feature", as.factor)
}
#allLociDML.features[[8]] <- allLociDML.features[[8]] %>%
#    setNames(c("contig.feat", "start.feat","end.feat","unknown"))
names(allLociDML.features) <- c("allLociDML.CDS","allLociDML.exon","allLociDML.gene","allLociDML.gene2kb","allLociDML.mRNA","allLociDML.TE","allLociDML.ASV","allLociDML.unknown")
allLociDML.features[["allLociDML.unknown"]] <- 
  allLociDML.features[["allLociDML.unknown"]] %>% 
  setNames(c("contig.allLoci", "start.allLoci", "end.allLoci", "unkown"))
allLociDML.features[["allLociDML.CDS"]] <- allLociDML.features[["allLociDML.CDS"]] %>% mutate_at("unknown2", as.character)
allLociDML.features[["allLociDML.ASV"]] <- allLociDML.features[["allLociDML.ASV"]] %>% mutate_at("unknown1", as.character)
allLociDML.features[["allLociDML.gene2kb"]]$feature <- "gene2kb"
allLociDML.features[["allLociDML.unknown"]]$feature <- "unknown"

allLociDML.features.df <- bind_rows(allLociDML.features[-7]) #don't include alternative splice variants or unknown 
print(allLociDML.summary <- table(allLociDML.features.df[c("feature")])) #Note, "similarity"= transposable elements 

save(allLociDML.summary, file="../analyses/DMLs/R-objects/allLociDML.summary") 
```

### Where are MACAU background loci located? 

```{r, message=FALSE, warning=FALSE}
allLociMACAUfiles <- c("AllLociMACAU-CDS.bed", "AllLociMACAU-exon.bed", "AllLociMACAU-gene.bed",  "AllLociMACAU-gene2kb.bed", "AllLociMACAU-mRNA.bed", "AllLociMACAU-TE.bed", "AllLociMACAU-ASV.bed", "AllLociMACAU-unknown.bed")
allLociMACAU.features <- list()
for (i in c(1:8)) {
  allLociMACAU.features[[i]] <- read_delim(here::here("analyses","macau", allLociMACAUfiles[i]), delim = '\t', col_names = FALSE) %>% as_tibble()}
for (i in 1:7) {
  allLociMACAU.features[[i]] <- allLociMACAU.features[[i]] %>%
    setNames(c("contig.allLoci","start.allLoci","end.allLoci", "contig.feat", "source.feat","feature","start.feat","end.feat","unknown1","strand","unknown2","attribute")) %>%
mutate(ID=str_extract(attribute, "ID=(.*?);"),
       Parent=str_extract(attribute, "Parent=(.*?);"),
       Name=str_extract(attribute, "Name=(.*?);"),
       Alias=str_extract(attribute, "Alias=(.*?);"),
       AED=str_extract(attribute, "AED=(.*?);"),
       eAED=str_extract(attribute, "eAED=(.*?);"),
       Note=str_extract(attribute, "Note=(.*?);"),
       Ontology_term=str_extract(attribute, "Ontology_term=(.*?);"),
       Dbxref=str_extract(attribute, "Dbxref=(.*?);")
       ) %>%
mutate_at("feature", as.factor)
}
names(allLociMACAU.features) <- c("allLociMACAU.CDS","allLociMACAU.exon","allLociMACAU.gene","allLociMACAU.gene2kb","allLociMACAU.mRNA","allLociMACAU.TE","allLociMACAU.ASV","allLociMACAU.unknown")
allLociMACAU.features[["allLociMACAU.unknown"]] <- 
  allLociMACAU.features[["allLociMACAU.unknown"]] %>% 
  setNames(c("contig.allLoci", "start.allLoci", "end.allLoci"))
allLociMACAU.features[["allLociMACAU.CDS"]] <- allLociMACAU.features[["allLociMACAU.CDS"]] %>% mutate_at("unknown2", as.character)
allLociMACAU.features[["allLociMACAU.ASV"]] <- allLociMACAU.features[["allLociMACAU.ASV"]] %>% mutate_at("unknown1", as.character)
allLociMACAU.features[["allLociMACAU.gene2kb"]]$feature <- "gene2kb"
allLociMACAU.features[["allLociMACAU.unknown"]]$feature <- "unknown"
allLociMACAU.features.df <- bind_rows(allLociMACAU.features[-7]) #don't include alternative splice variants or unknown 
print(allLociMACAU.summary <- table(allLociMACAU.features.df[c("feature")])) #Note, "similartiy"= transposable elements 

save(allLociMACAU.summary, file="../analyses/macau/R-objects/allLociMACAU.summary")
```

## Summary barplot showing where DMLs, MACAU and All Loci are located (relative to the total number assessed for each groups)

```{r, message=FALSE, warning=FALSE}
load("../analyses/DMLs/R-objects/dml25") # load df for no. of DMLs (including those not annotated to known features)
load("../analyses/DMLs/R-objects/mydiff.all") # load df for total no. of loci (which were analyzed for DML) 
load("../analyses/macau/R-objects/macau.FDR.length") #load df for total no. of MACAU loci analyzed & no. of sign. loci 

# Add row to summary dataframe with total number of loci in DMLs, all loci 

# Create single summary df showing where DML loci and all loci analyzed for DML are located 
loci.locations <-  merge(x=merge(x=merge(x=melt(DML.summary, varnames = "feature", value.name = "DML"), 
      y=melt(macau.summary, varnames = "feature", value.name = "macau"),
      by="feature", all=TRUE), 
      y=melt(allLociDML.summary, varnames = "feature", value.name = "allLociDML"),
      by="feature", all=TRUE),
      y=melt(allLociMACAU.summary, varnames = "feature", value.name = "allLociMACAU"),
      by="feature", all=TRUE) %>% 
  mutate(feature = as.character(feature)) %>%
  mutate(DML=as.numeric(DML), allLociDML=as.numeric(allLociDML))
#    DML=as.numeric(DML), allLociDML=as.numeric(allLociDML)) %>%

loci.locations <- loci.locations  %>%
  rbind(c("geneflank2kb",
          loci.locations[loci.locations$feature=="gene2kb","DML"]-loci.locations[loci.locations$feature=="gene","DML"],
          loci.locations[loci.locations$feature=="gene2kb","macau"]-loci.locations[loci.locations$feature=="gene","macau"],
          loci.locations[loci.locations$feature=="gene2kb","allLociDML"]-loci.locations[loci.locations$feature=="gene","allLociDML"],
          loci.locations[loci.locations$feature=="gene2kb","allLociMACAU"]-loci.locations[loci.locations$feature=="gene","allLociMACAU"])) %>%
  rbind(c("all", as.numeric(nrow(dml25)), as.numeric(nrow(subset(macau.FDR.length, significant=="TRUE"))), as.numeric(nrow(mydiff.all)), as.numeric(nrow(macau.FDR.length)))) %>%
  mutate_at(vars(-feature), funs(as.numeric)) %>%
  mutate(DMLperc=DML/nrow(dml25), 
         MACAUperc=macau/nrow(subset(macau.FDR.length, significant=="TRUE")),
         AllDMLperc=allLociDML/nrow(mydiff.all),
         AllMACAUperc=allLociMACAU/nrow(macau.FDR.length))
save(loci.locations, file="../analyses/methylation-filtered/R-objects/loci.locations")

loci.locations.long <- cbind(melt(loci.locations[,c("feature", "DML", "macau", "allLociDML", "allLociMACAU")], 
                                  variable.name = "analysis", value.name = "count"),
                             melt(loci.locations[,c("feature", "DMLperc", "MACAUperc", "AllDMLperc", "AllMACAUperc")], 
                                  variable.name = "analysis", value.name = "percent"))[,c(1,2,3,6)] %>%
   mutate(analysis=fct_relevel(analysis, c("DML", "allLociDML", "macau", "allLociMACAU")),
         feature=as.factor(str_replace(feature, "similarity", "TE")))
save(loci.locations.long, file="../analyses/methylation-filtered/R-objects/loci.locations.long")
```


```{r}
ggplot(data=subset(loci.locations.long, feature=="gene" | feature=="geneflank2kb" | feature=="exon" | feature=="TE" | feature=="unknown"), aes(x=analysis, y=percent, fill=feature, label=percent(percent, accuracy = 0.1))) +  #prettyNum(count, big.mark = ",")
#  geom_bar(stat="identity", width = .5) +
  geom_bar(position="fill", stat="identity", width=0.5) +
  labs(y="% of Loci Assessed", x=NULL) + 
  scale_fill_manual(name = "Loci Location", labels = c("Exon", "Gene Body", "Gene Flanking Regions (+/-2kb)", "Transposable Elements", "Unknown Regions"), 
                    values=c("#a6cee3", "#1f78b4", "#b2df8a","#33a02c", "#fb9a99")) + 
  ggtitle("Locations of DML and size-associated (MACAU) loci in genome\n with background loci locations") +
  theme_minimal() + geom_text(size = 3, position = position_fill(vjust = 0.5)) + 
  scale_x_discrete(labels=c("DML" = "DMLs",
                            "allLociDML" = "DML Background\nLoci",
                            "macau" = "Size-Associated\nLoci (MACAU)",
                            "allLociMACAU" = "MACAU\nBackground\nLoci"))
```

## Extract GO terms 

```{r}
###### ====== DML loci GO terms  

# Count string lengths in the Ontology_term column and check out - to figure out max # GO terms 
#data.frame(ontology=DML.features.df$Ontology_term,chr=apply(DML.features.df[,"Ontology_term"],2,nchar)) # %>% View()

DML.GO <- DML.features.df[c("contig.dml", "start.dml", "end.dml", "feature", "start.feat", "end.feat", "Note", "Ontology_term")] %>%
  distinct(contig.dml, start.dml, Note, Ontology_term, .keep_all = TRUE) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern="Ontology_term=",replacement = "")) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern=";",replacement = "")) %>%
  separate(Ontology_term, sep=",", into=paste("GO", 1:11, sep="_")) %>%
  pivot_longer(cols=c("GO_1","GO_2","GO_3","GO_4","GO_5","GO_6","GO_7","GO_8","GO_9","GO_10","GO_11"), names_to = "GO_number", values_to = "GO_term") %>%
  select(-GO_number) %>%
  filter(!is.na(Note) & !is.na(GO_term)) 

write_csv(DML.features.df, path = here::here("analyses/", "DMLs", "DML.features.txt")) #write out df with all DML features 
write(DML.GO$GO_term, file = here::here("analyses/", "DMLs", "DML.GO.txt"))
write.table(subset(DML.features.df, feature=="gene"), here::here("analyses/", "DMLs", "DML.geneinfo.txt"), sep = "\t") #write out gene info for DMLs to associate with UniprotIDs

###### ====== MACAU loci GO terms  

# Count string lengths in the Ontology_term column and check out - to figure out max # GO terms 
#data.frame(ontology=macau.features.df$Ontology_term,chr=apply(macau.features.df[,"Ontology_term"],2,nchar)) #%>% View()

macau.GO <- macau.features.df[c("contig.macau", "start.macau", "end.macau", "feature", "start.feat", "end.feat", "Note", "Ontology_term")] %>%
  distinct(contig.macau, start.macau, Note, Ontology_term, .keep_all = TRUE) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern="Ontology_term=",replacement = "")) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern=";",replacement = "")) %>%
  separate(Ontology_term, sep=",", into=paste("GO", 1:6, sep="_")) %>%
  pivot_longer(cols=c("GO_1","GO_2","GO_3","GO_4","GO_5","GO_6"), names_to = "GO_number", values_to = "GO_term") %>%
  select(-GO_number) %>%
  filter(!is.na(Note) & !is.na(GO_term))

write_csv(macau.features.df, path = here::here("analyses/", "macau/", "macau.features.csv")) #write out df with all MACAU features 
write(macau.GO$GO_term, file = here::here("analyses/", "macau/", "macau.GO.txt")) #write out df with just GO terms 
write.table(subset(macau.features.df, feature=="gene"), here::here("analyses/", "macau/", "MACAU.geneinfo.txt"), sep = "\t") #write out gene info for MACAU loci to associate with UniprotIDs

###### ====== DML Background loci GO terms  

# Count string lengths in the Ontology_term column and check out - to figure out max # GO terms 
#data.frame(ontology=allLociDML.features.df$Ontology_term,chr=apply(allLociDML.features.df[,"Ontology_term"],2,nchar)) # %>% View()

allLociDML.GO <- allLociDML.features.df[c("contig.allLoci", "start.allLoci", "end.allLoci", "feature", "start.feat", "end.feat", "Note", "Ontology_term")] %>%
  distinct(contig.allLoci, start.allLoci, Note, Ontology_term, .keep_all = TRUE) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern="Ontology_term=",replacement = "")) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern=";",replacement = "")) %>%
  separate(Ontology_term, sep=",", into=paste("GO", 1:11, sep="_")) %>%
  pivot_longer(cols=c("GO_1","GO_2","GO_3","GO_4","GO_5","GO_6","GO_7","GO_8","GO_9","GO_10","GO_11"), names_to = "GO_number", values_to = "GO_term") %>%
  select(-GO_number) %>%
  filter(!is.na(Note) & !is.na(GO_term)) 
write_csv(allLociDML.features.df, path = here::here("analyses/", "DMLs", "allLociDML.features.txt")) #write out df with all loci features 
write(allLociDML.GO$GO_term, file = here::here("analyses/", "DMLs", "allLociDML.GO.txt"))
write.table(subset(allLociDML.features.df, feature=="gene"), here::here("analyses/", "DMLs", "allLociDML.geneinfo.txt"), sep = "\t") #write out gene info for All loci fed into DMLs to associate with UniprotIDs

#save all loci feature df object to file, to use in notebook #12 
save(allLociDML.features.df, file=here::here("analyses", "DMLs", "R-objects", "allLociDML.features.df")) 

###### ====== MACAU Background loci GO terms  

# Count string lengths in the Ontology_term column and check out - to figure out max # GO terms 
#data.frame(ontology=allLociMACAU.features.df$Ontology_term,chr=apply(allLociMACAU.features.df[,"Ontology_term"],2,nchar))  #%>% View()

allLociMACAU.GO <- allLociMACAU.features.df[c("contig.allLoci", "start.allLoci", "end.allLoci", "feature", "start.feat", "end.feat", "Note", "Ontology_term")] %>%
  distinct(contig.allLoci, start.allLoci, Note, Ontology_term, .keep_all = TRUE) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern="Ontology_term=",replacement = "")) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern=";",replacement = "")) %>%
  separate(Ontology_term, sep=",", into=paste("GO", 1:11, sep="_")) %>%
  pivot_longer(cols=c("GO_1","GO_2","GO_3","GO_4","GO_5","GO_6","GO_7","GO_8","GO_9","GO_10","GO_11"), names_to = "GO_number", values_to = "GO_term") %>%
  select(-GO_number) %>%
  filter(!is.na(Note) & !is.na(GO_term)) 
write_csv(allLociMACAU.features.df, path = here::here("analyses/", "macau", "allLociMACAU.features.txt")) #write out df with all loci features 
write(allLociMACAU.GO$GO_term, file = here::here("analyses/", "macau", "allLociMACAU.GO.txt"))
write.table(subset(allLociMACAU.features.df, feature=="gene"), here::here("analyses/", "macau", "allLociMACAU.geneinfo.txt"), sep = "\t") #write out gene info for All loci fed into MACAU to associate with UniprotIDs

#save all loci feature df object to file, to use in notebook #12 
save(allLociMACAU.features.df, file=here::here("analyses", "macau", "R-objects", "allLociMACAU.features.df")) 
```


# BONEYARD

## Create separate dataframes with annotation information for all hypomethylated loci in Hood Canal & South Sound, separately 

```{r, eval=FALSE}
hypo_HC_genes <-
  DML.features.df %>%
  filter(contig.dml %in% hypo_HC$chr &
         start.dml %in% 1+hypo_HC$start &
         feature == "gene" &
         Note != "Note=Protein of unknown function;") %>%
  distinct(contig.dml, start.dml, Note, Ontology_term, .keep_all = TRUE)

hypo_HC_GO <- hypo_HC_genes %>%
mutate(Ontology_term = str_replace(Ontology_term, pattern="Ontology_term=",replacement = "")) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern=";",replacement = "")) %>%
  separate(Ontology_term, sep=",", into=paste("GO", 1:3, sep="_")) %>%
  pivot_longer(cols=c("GO_1","GO_2","GO_3"), names_to = "GO_number", values_to = "GO_term") %>%
  select(-GO_number) %>%
  filter(!is.na(Note) & !is.na(GO_term))
write(hypo_HC_GO$GO_term, file = here::here("analyses/", "hypo_HC_GO.txt"))


hypo_SS_genes <- DML.features.df %>%
  filter(contig.dml %in% hypo_SS$chr &
         start.dml %in% 1+hypo_SS$start &
           feature == "gene" &
           Note != "Note=Protein of unknown function;") %>%
  distinct(contig.dml, start.dml, Note, Ontology_term, .keep_all = TRUE)

hypo_SS_GO <- hypo_SS_genes %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern="Ontology_term=",replacement = "")) %>%
  mutate(Ontology_term = str_replace(Ontology_term, pattern=";",replacement = "")) %>%
  separate(Ontology_term, sep=",", into=paste("GO", 1:5, sep="_")) %>%
  pivot_longer(cols=c("GO_1","GO_2","GO_3", "GO_4", "GO_5"), names_to = "GO_number", values_to = "GO_term") %>%
  select(-GO_number) %>%
  filter(!is.na(Note) & !is.na(GO_term))
write(hypo_HC_GO$GO_term, file = here::here("analyses/", "hypo_SS_GO.txt"))


# Barplots of differentially methylated loci that are located in known genes

hypo_HC %>%
   filter(chr %in% hypo_HC_genes$contig.dml &
         start %in% hypo_HC_genes$start.dml-1) %>%
ggplot(aes(x = population, y = mean_percMeth, fill = population,
                         label=paste0(round(mean_percMeth, digits = 2), "%"))) +
      geom_bar(stat="identity", width = 0.5) + ylim(0,110) +
      geom_pointrange(aes(ymin=mean_percMeth,
                        ymax=mean_percMeth+sd_percMeth, width=0.15)) +
      geom_text(size=3, vjust=-0.5, hjust=1.1) +
      theme_minimal() + xlab("Population") + ylab("% Methylation") +
  scale_y_continuous(breaks=c(0,25,50,75,100)) +
    scale_fill_manual(values=c("firebrick3","dodgerblue3")) +
   facet_wrap_paginate(~chr + start, nrow=2, ncol=3, page=2)


hypo_SS %>%
   filter(chr %in% hypo_SS_genes$contig.dml &
         start %in% hypo_SS_genes$start.dml-1) %>%
ggplot(aes(x = population, y = mean_percMeth, fill = population,
                         label=paste0(round(mean_percMeth, digits = 2), "%"))) +
      geom_bar(stat="identity", width = 0.5) + ylim(0,110) +
      geom_pointrange(aes(ymin=mean_percMeth,
                        ymax=mean_percMeth+sd_percMeth, width=0.15)) +
      geom_text(size=3, vjust=-0.5, hjust=1.1) +
      theme_minimal() + xlab("Population") + ylab("% Methylation") +
  scale_y_continuous(breaks=c(0,25,50,75,100)) +
    scale_fill_manual(values=c("firebrick3","dodgerblue3")) +
   facet_wrap_paginate(~chr + start, nrow=2, ncol=3, page=2)
```


```{bash, eval=FALSE}
echo "Loci differentially methylated between SS and HC populations:"
cat "../analyses/dml25.bed" | wc -l 

echo "Loci that overlap with genes:"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" | wc -l

echo "Loci that overlap with gene regions (genes +/- 2kb):"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" | wc -l

echo "Loci that overlap with exons:"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" | wc -l

echo "Loci that overlap with coding sequences:"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" | wc -l

echo "Loci that overlap with mRNA:"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" | wc -l

echo "Loci that overlap with transposable elements:"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" | wc -l

echo "Loci that overlap with alternative splice variants:"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l

echo "Loci that do not overlap with known features:"
bedtools intersect -u -a "../analyses/dml25.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l
```


```{bash, eval=FALSE}
echo "genes" 
bedtools intersect -u -a "../analyses/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" | wc -l
echo "gene regions (+/- 2kb)" 
bedtools intersect -u -a "../analyses/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" | wc -l
echo "exon" 
bedtools intersect -u -a "../analyses/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" | wc -l
echo "CDS" 
bedtools intersect -u -a "../analyses/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" | wc -l
echo "mRNA"
bedtools intersect -u -a "../analyses/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" | wc -l
echo "TE" 
bedtools intersect -u -a "../analyses/mydiff-all.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" | wc -l
echo "ASV" 
bedtools intersect -u -a "../analyses/mydiff-all.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l
echo "unknown" 
bedtools intersect -v -a "../analyses/mydiff-all.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l
```


```{bash, eval=FALSE}
echo "Total methylated loci:" 
cat ../analyses/macau/macau-all-loci.bed | wc -l

echo "Loci associated with shell length (MACAU):"
cat "../analyses/macau/macau.sign.length.perc.meth.bed" | wc -l 

echo "Loci that overlap with genes:"
bedtools intersect -u -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" | wc -l

echo "Loci that overlap with gene regions (+/- 2kb):"
bedtools intersect -u -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.gene.2kbslop.gff" | wc -l

echo "Loci that overlap with exons:"
bedtools intersect -u -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.exon.gff" | wc -l

echo "Loci that overlap with coding sequences:"
bedtools intersect -u -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.CDS.gff" | wc -l

echo "Loci that overlap with mRNA:"
bedtools intersect -u -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.mRNA.gff" | wc -l

echo "Loci that overlap with transposable elements:"
bedtools intersect -u -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081_TE-Cg.gff" | wc -l

echo "Loci that overlap with alternative splice variants:"
bedtools intersect -u -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l

echo "Loci that do not overlap with known features:"
bedtools intersect -v -a "../analyses/macau/macau.sign.length.perc.meth.bed" -b "../genome-features/Olurida_v081-20190709.gene.gff" "../genome-features/Olurida_v081-20190709.exon.gff" "../genome-features/Olurida_v081-20190709.CDS.gff" "../genome-features/Olurida_v081-20190709.mRNA.gff" "../genome-features/Olurida_v081_TE-Cg.gff" "../genome-features/20190709-Olurida_v081.stringtie.gtf" | wc -l
```
