---
title: "06-Analyzing MACAU results-rev2"
author: "Laura H Spencer"
date: "1/29/2020"
output: html_document
---
## MACAU results - visualize and FDR correct  

Ran MACAU in a Jupyter Notebook. See [05-MACAU.ipynb](https://github.com/sr320/paper-oly-mbdbs-gen/blob/master/code/05-MACAU.ipynb)

In this notebook I assess patterns in methylated loci that MACAU identified as influence by oyster phenotype, aka oyster shell length. I ran MACAU twice using two sets of count files: 

1) Contigs retained if 75% of samples (14 out of 18) had 10x coverage
- Total count data: analyses/counts.tot.destrand.10x75
- Methylated count data: analyses/counts.meth.destrand.10x75

2) Contigs retained if 75% of samples (14 out of 18) had 5x coverage

- Total count data: analyses/counts.tot.destrand.5x75.txt
- Methylated count data: analyses/counts.meth.destrand.5x75.txt

See my Jupyter notebook on these runs: [code/05-MACAU.ipynb](https://github.com/sr320/paper-oly-mbdbs-gen/blob/master/code/05-MACAU.ipynb)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries 

```{r, message=FALSE, warning=FALSE, results=FALSE}
list.of.packages <- c("vegan", "dplyr", "factoextra", "FactoMineR", "tidyverse", "tibble", "reshape2", "gplots", "qvalue", "cluster", "here", "gridExtra") #add new libraries here 
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load all libraries 
lapply(list.of.packages, FUN = function(X) {
  do.call("require", list(X)) 
})
```

### Read in count data 
Use dataframe that has the siteID, and original contig, start #, and end # info 

```{r}
head(counts.meth.destrand.10x75 <- read.table(file = here::here("analyses", "counts.meth.destrand.10x75.txt"), sep = "\t", header=T, stringsAsFactors = F))
head(counts.tot.destrand.10x75 <- read.table(file = here::here("analyses", "counts.tot.destrand.10x75.txt"), sep = "\t", header=T, stringsAsFactors = F))
```

### Read in dataframe with MACAU assocation results 

```{r}
macau.10x75 <- read.table(file=here::here("analyses","macau","output","20200107-macau-10x75.assoc.txt"), sep = "\t", stringsAsFactors = F)
colnames(macau.10x75) <- macau.10x75[1,]
macau.10x75 <- macau.10x75[-1, ] 
macau.10x75$pvalue <- as.numeric(macau.10x75$pvalue)
hist(macau.10x75$pvalue)

# Read dataframes in if needed (object created in noteboook 04)
#counts.tot.destrand.10x75 <- read_delim(here::here("analyses", "counts.tot.destrand.10x75.txt"), delim = "\t")
summary(macau.10x75$id == counts.tot.destrand.10x75$siteID) #confirm that macau results and original destranded dataframe are in same order 

# bind macau results with df that has contig id and loci information 
macau.10x75.c <- cbind(counts.tot.destrand.10x75[,1],macau.10x75) %>%
   separate(`counts.tot.destrand.10x75[, 1]`, into = c("contig", "locus"), sep = "_")

write.table(macau.10x75.c, file = here::here("analyses", "macau", "macau.10x75.c.tab"), sep = "\t", row.names = FALSE,  quote = FALSE)
```

### Correct for multiple comparisons 

From Mac's 2019 salmon paper, [doi:10.3390/genes10050356](10.3390/genes10050356)  
"Multiple hypothesis testing was performed on P-values extracted from the MACAU output for each CG site using the false discovery rate (FDR) approach used in the R package qvalue [45]. We considered a CG site to be differentially methylated (i.e., differentially methylated cytosine (DMC) if it passed a 10% FDR threshold, consistent with [46]."

```{r}
hist(macau.10x75.c$pvalue) # p-value distribution looks good (i.e. not u-shaped)
macau.qvalue.10x75 <- qvalue(p = macau.10x75.c$pvalue, fdr.level = 0.1) # create qvalue object, setting FDR=10%
summary(macau.qvalue.10x75) 
100*(1- macau.qvalue.10x75$pi0) #% of all loci in df as truely sign.  
hist(macau.qvalue.10x75)

#if one wants to estimate the false discovery rate when calling all p-values ≤ 0.01 significant: 
max(macau.qvalue.10x75$qvalues[macau.qvalue.10x75$pvalues <= 0.01])

# The main purpose of the upper-left plot is to gauge the reliability of the π0 estimate, where the estimated π0 is plotted versus the tuning parameter λ. The remaining plots show how many tests are significant, as well as how many false positives to expect for each q-value cut-off.
plot(macau.qvalue.10x75)

# Given our FDR=10% level, how many are significant? 
summary(macau.qvalue.10x75$significant) #219 sign. (# TRUE)

# Add qvalue results to macau df 
macau.FDR.10x75 <- cbind(macau.10x75.c, do.call(cbind.data.frame, macau.qvalue.10x75[c("qvalues", "pvalues", "significant")]))

#plot macau p-values against pvalues in the qvalue results - should all fall on 1:1 line (double check that pvalues are in same order) - 
plot(macau.FDR.10x75$pvalue, macau.FDR.10x75$pvalues) # looks good! 
```

### Extract IDs for loci where MACAU indicates DML, FDR test (at 10%) deemed loci significantt 

```{r}
head(macau.sign.10x75 <- subset(macau.FDR.10x75, significant=="TRUE"), n=20)
nrow(macau.sign.10x75) #72
```

### Merge MACAU + FDR results with raw count data 
**This constitutes a master dataframe** 

```{r}
# merge macau results with count dataframe, save to file 
head(macau.sign.meth.counts.10x75 <- merge(x=macau.sign.10x75, y=subset(counts.meth.destrand.10x75, siteID %in% macau.sign.10x75$id), by.x="id", by.y="siteID"))
head(macau.sign.tot.counts.10x75 <- merge(x=macau.sign.10x75, y=subset(counts.tot.destrand.10x75, siteID %in% macau.sign.10x75$id), by.x="id", by.y="siteID"))

write.table(macau.sign.meth.counts.10x75, file = here::here("analyses", "combined-macau-FDR-meth-counts.10x75.txt"), sep = "\t", row.names = FALSE,  quote = FALSE) 
write.table(macau.sign.tot.counts.10x75, file = here::here("analyses", "combined-macau-FDR-tot-counts.10x75.txt"), sep = "\t", row.names = FALSE,  quote = FALSE) 
```

### Create df where coverage counts <10 replaced with NA 

```{r}
head(macau.sign.tot.counts.10x75.na <- macau.sign.tot.counts.10x75 %>% 
  mutate_at(vars(contains('coverage')), funs(ifelse(. < 10, NA, .))))
```

### Create % methylated dataframe with sign. MACAU loci 

```{r}
# duplicate macau methylated count dataframe to make % methylated df for 10x coverage (all <10x cov will be "NA"")
macau.sign.perc.meth.10x75 <- macau.sign.meth.counts.10x75

# divide macau methylated counts by total counts 
macau.sign.perc.meth.10x75[,20:37] <- 100*(macau.sign.meth.counts.10x75[,20:37]/macau.sign.tot.counts.10x75.na[,20:37])
rownames(macau.sign.perc.meth.10x75) <- macau.sign.perc.meth.10x75[,1]
```

### Heatmap of % methylated, cluster analysis order, with NA for instances that loci coverage <10x 
Sample ordered from cluster analysis, where dendogram is constructed from Gower's distance matrix (lots of NA values), and Ward cluster method. 

---> mbd samples #1-9 = Hood Canal population  
---> mbd samples #10-18 = South Sound population  

### Heatmap of % methylated, cluster analysis order. 

- Only loci with 10x coverage in color, less than 10x coverage in white.  
- Sample ordered from cluster analysis, where dendogram is constructed from correlation matrix, and Ward cluster method.  

```{r}
heatmap.2(as.matrix(macau.sign.perc.meth.10x75[20:37]), 
          main = "MACAU id'd % methylated\n10x cov\nOrdered by cluster analysis", 
          xlab = NA, ylab = "loci",  key.ylab = NA, key.title = "% methylated",
          hclustfun = function(x) hclust(x, method="ward.D"), distfunc <- function(x) daisy(x,metric="correlation"), trace="none", dendrogram = "column", na.color = "white",
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = c(rep("firebrick3", times=9), rep("dodgerblue3", times=9)))

# Save same heatmap to file 
pdf(file ="../analyses/macau/macau-heatmap.10x75.cluster.pdf")
par(cex.main=.8, oma=c(5,1,2,5))
heatmap.2(as.matrix(macau.sign.perc.meth.10x75[20:37]), 
          main = "MACAU id'd % methylated\n10x cov\nOrdered by cluster analysis", 
          xlab = NA, ylab = "loci",  key.ylab = NA, key.title = "% methylated",
          hclustfun = function(x) hclust(x, method="ward.D"), distfunc <- function(x) daisy(x,metric="correlation"), trace="none", dendrogram = "column", na.color = "white",
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = c(rep("firebrick3", times=9), rep("dodgerblue3", times=9)))
dev.off()
```

### Heatmap of % methylated, tree order, with NA for instances that loci coverage <10x  
No cluster analysis, instead samples are presented in the same order as the MethylKit tree. 

```{r}
tree.order <- paste("numCs",c(15,17,12,10,11,13,18,14,16,8,1,6,9,2,4,5,3,7), sep="")
pop.colors.tree <- c(rep("dodgerblue3", times=9), rep("firebrick3", times=9))

heatmap.2(as.matrix(macau.sign.perc.meth.10x75[tree.order]), main = "MACAU id'd % methylated\nwhite=<10x cov\nOrdered by MethylKit tree", key.ylab = NA, key.title = "% methylated",
          xlab = NA, ylab = "loci", na.color = "white",trace="none", dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = pop.colors.tree)

# Save same heatmap to file 
pdf(file ="../analyses/macau/macau-heatmap.10x75.tree.pdf")
par(cex.main=.8, oma=c(5,1,2,5))
heatmap.2(as.matrix(macau.sign.perc.meth.10x75[tree.order]), main = "MACAU id'd % methylated\nwhite=<10x cov\nOrdered by MethylKit tree", key.ylab = NA, key.title = "% methylated",
          xlab = NA, ylab = "loci", na.color = "white",trace="none", dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = pop.colors.tree)
dev.off()
```

### Heatmap of % methylated, ordered by shell length 
No cluster analysis, instead samples are presented in the same order as the MethylKit tree. 

### Prepare size data to plot in same order as this heat map 
This is laborious code with some room for error - needs fixing  

```{r}
size <- read.csv(file = here::here("data","mbd_size.csv"), header=T, sep = "\t")
key <- read.csv(file=here::here("data","sample-key.csv"))
size.macau <- cbind(x=c(rep(1,times=18)), y=merge(y=size, x=key, by.y="Sample", by.x="SAMPLE")) %>%
  mutate(population=as.factor(c(rep("HC", times=9), rep("SS",  times=9)))) %>%
  mutate(sample=paste("numCs", y.MBD.FILENAME, sep="")) %>%
  arrange(y.Length)
mycols <- c("firebrick3", "dodgerblue3")
pop.colors.length <- c("dodgerblue3", rep("firebrick3", times=7), "dodgerblue3", "dodgerblue3", "dodgerblue3", "firebrick3", "dodgerblue3", "firebrick3", rep("dodgerblue3", times=5))

heatmap.2(as.matrix(macau.sign.perc.meth.10x75[!rowSums(is.na(macau.sign.perc.meth.10x75)) > 0.25*18,size.macau$sample]), 
          main = "MACAU id'd % methylation\nOrdered by shell length\nwhite=<10x cov", xlab = NA, ylab = "loci", 
          key.ylab = NA, key.title = "% methylated",
          trace="none",na.color = "white",dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = pop.colors.length)
barplot(height = size.macau$y.Length,  names.arg = size.macau$sample, col = mycols[size.macau$pop], cex.names =  0.8,
        main = "Shell Length (mm)", las=2, ylab = "Shell Length (mm)")

# Save same plots to file 
pdf(file ="../analyses/macau/macau-heatmap.10x75.length.pdf")
par(cex.main=.8, oma=c(5,1,2,5))
heatmap.2(as.matrix(macau.sign.perc.meth.10x75[!rowSums(is.na(macau.sign.perc.meth.10x75)) > 0.25*18,size.macau$sample]), 
          main = "MACAU id'd % methylation\nOrdered by shell length\nwhite=<10x cov", xlab = NA, ylab = "loci", 
          key.ylab = NA, key.title = "% methylated",
          trace="none",na.color = "white",dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = pop.colors.length)
par(oma=c(2,4,5,2))
barplot(height = size.macau$y.Length,  names.arg = size.macau$sample, col = mycols[size.macau$pop], cex.names =  0.8,
        main = "Shell Length (mm)", las=2, ylab = "Shell Length (mm)")
dev.off()
```


### Create BED files for MACAU id'd loci with 10x coverage for 1) any samples, 2) all samples 

```{r}
macau.sign.perc.meth.10x75 %>% 
  mutate(end = locus) %>% 
  rename(start = locus) %>%
  dplyr::select(c("contig", "start", "end")) %>%
write_delim(here::here("analyses","macau", "macau.sign.perc.meth.10x75.bed"),  delim = '\t', col_names = FALSE)

# All loci fed into/from macau 
macau.10x75.c %>%
  mutate(end = locus) %>% 
  rename(start = locus) %>%
  dplyr::select(c("contig", "start", "end")) %>%
write_delim(here::here("analyses","macau", "macau-all-loci.10x75.bed"),  delim = '\t', col_names = FALSE)
```

# Re-do analysis with 5x75 

That is, MACAU results which analyzed count data for loci where 75% of samples had **5x coverage** or greater. 

```{r}
head(counts.meth.destrand.5x75 <- read.table(file = here::here("analyses", "counts.meth.destrand.5x75.txt"), sep = "\t", header=T, stringsAsFactors = F))
head(counts.tot.destrand.5x75 <- read.table(file = here::here("analyses", "counts.tot.destrand.5x75.txt"), sep = "\t", header=T, stringsAsFactors = F))

macau.5x75 <- read.table(file=here::here("analyses","macau","output","20200108-macau-5x75.assoc.txt"), sep = "\t", stringsAsFactors = F)
colnames(macau.5x75) <- macau.5x75[1,]
macau.5x75 <- macau.5x75[-1, ] 
macau.5x75$pvalue <- as.numeric(macau.5x75$pvalue)
hist(macau.5x75$pvalue)

summary(macau.5x75$id == counts.tot.destrand.5x75$siteID) #confirm that macau results and original destranded dataframe are in same order 
macau.5x75.c <- cbind(counts.tot.destrand.5x75[,1],macau.5x75) %>%
   separate(`counts.tot.destrand.5x75[, 1]`, into = c("contig", "locus"), sep = "_")
write.table(macau.5x75.c, file = here::here("analyses", "macau", "macau.5x75.c.tab"), sep = "\t", row.names = FALSE,  quote = FALSE)
```

### Correct for multiple comparisons 

```{r}
hist(macau.5x75.c$pvalue) # p-value distribution looks good (i.e. not u-shaped)
macau.qvalue.5x75 <- qvalue(p = macau.5x75.c$pvalue, fdr.level = 0.1) # create qvalue object, setting FDR=10%
summary(macau.qvalue.5x75) 
100*(1- macau.qvalue.5x75$pi0) #% of all loci in df as truely sign.  
hist(macau.qvalue.5x75)
max(macau.qvalue.5x75$qvalues[macau.qvalue.5x75$pvalues <= 0.01])
plot(macau.qvalue.5x75)
summary(macau.qvalue.5x75$significant) #219 sign. (# TRUE)
macau.FDR.5x75 <- cbind(macau.5x75.c, do.call(cbind.data.frame, macau.qvalue.5x75[c("qvalues", "pvalues", "significant")]))
plot(macau.FDR.5x75$pvalue, macau.FDR.5x75$pvalues) # looks good! 
```

### Extract IDs for loci where MACAU indicates DML, FDR test (at 10%) deemed loci significantt 
### Merge MACAU + FDR results with raw count data 
**This constitutes a master dataframe** 

```{r}
head(macau.sign.5x75 <- subset(macau.FDR.5x75, significant=="TRUE"), n=20)
nrow(macau.sign.5x75) #146
head(macau.sign.meth.counts.5x75 <- merge(x=macau.sign.5x75, y=subset(counts.meth.destrand.5x75, siteID %in% macau.sign.5x75$id), by.x="id", by.y="siteID"))
head(macau.sign.tot.counts.5x75 <- merge(x=macau.sign.5x75, y=subset(counts.tot.destrand.5x75, siteID %in% macau.sign.5x75$id), by.x="id", by.y="siteID"))
write.table(macau.sign.meth.counts.5x75, file = here::here("analyses", "combined-macau-FDR-meth-counts.5x75.txt"), sep = "\t", row.names = FALSE,  quote = FALSE) 
write.table(macau.sign.tot.counts.5x75, file = here::here("analyses", "combined-macau-FDR-tot-counts.5x75.txt"), sep = "\t", row.names = FALSE,  quote = FALSE) 
```

### Create df where coverage counts <5 replaced with NA, create % methylated dataframe with sign. MACAU loci 

```{r}
macau.sign.tot.counts.5x75.na <- macau.sign.tot.counts.5x75 %>% 
  mutate_at(vars(contains('coverage')), funs(ifelse(. < 5, NA, .)))
macau.sign.perc.meth.5x75 <- macau.sign.meth.counts.5x75
macau.sign.perc.meth.5x75[,20:37] <- 100*(macau.sign.meth.counts.5x75[,20:37]/macau.sign.tot.counts.5x75.na[,20:37])
rownames(macau.sign.perc.meth.5x75) <- macau.sign.perc.meth.5x75[,1]
```

### Heatmaps (not saved to file)

```{r}
par(cex.main=.8)
heatmap.2(as.matrix(macau.sign.perc.meth.5x75[20:37]), 
          main = "MACAU id'd % methylated\n5x cov\nOrdered by cluster analysis", 
          xlab = "mbd sample", ylab = "loci",  
          hclustfun = function(x) hclust(x, method="ward.D"), distfunc <- function(x) daisy(x,metric="correlation"), trace="none", dendrogram = "column", na.color = "white",
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = c(rep("firebrick3", times=9), rep("dodgerblue3", times=9)))

par(cex.main=.8)
heatmap.2(as.matrix(macau.sign.perc.meth.5x75[tree.order]), main = "MACAU id'd % methylated\ngray=<5x cov\nOrdered by MethylKit tree", 
          xlab = "mbd sample", ylab = "loci", na.color = "white",trace="none", dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = pop.colors.tree)

par(cex.main=.8)
heatmap.2(as.matrix(macau.sign.perc.meth.5x75[!rowSums(is.na(macau.sign.perc.meth.5x75)) > 0.25*18,size.macau$sample]), 
          main = "MACAU id'd % methylation\nFDR=10%, loci cov=5x (75% of samples)\nOrdered by Length", xlab = "mbd sample", ylab = "loci", 
          trace="none", dendrogram = "none", Rowv=FALSE, Colv = FALSE,
          col = colorRampPalette(c('#08519c','#bdd7e7'))(30), colCol = pop.colors.length)
par(oma=c(0,8,0,0))
barplot(height = size.macau$y.Length,  names.arg = size.macau$sample, col = mycols[size.macau$pop], cex.names =  0.8,
        main = "Shell Length (mm)", las=2)
```

### Create BED files for 1) MACAU id'd loci with 5x coverage, and 2) all loci with 5x coverage  

```{r}
macau.sign.perc.meth.5x75 %>% 
  mutate(end = locus) %>% 
  rename(start = locus) %>%
  dplyr::select(c("contig", "start", "end")) %>%
write_delim(here::here("analyses","macau", "macau.sign.perc.meth.5x75.bed"),  delim = '\t', col_names = FALSE)

# All loci fed into/from macau 
macau.5x75.c %>%
  mutate(end = locus) %>% 
  rename(start = locus) %>%
  dplyr::select(c("contig", "start", "end")) %>%
write_delim(here::here("analyses","macau", "macau-all-loci.5x75.bed"),  delim = '\t', col_names = FALSE)
```
