---
title: "06-Analyzing MACAU results"
author: "Laura H Spencer"
date: "9/6/2019"
output: html_document
---
## Visualize MACAU results 

Ran MACAU in a Jupyter Notebook. See [05-MACAU.ipynb](https://github.com/sr320/paper-oly-mbdbs-gen/blob/master/code/05-MACAU.ipynb)

In this notebook I assess patterns in methylated loci that MACAU identified as influence by oyster phenotype, aka oyster shell length. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load libraries 

```{r, message=FALSE, warning=FALSE, results=FALSE}
list.of.packages <- c("vegan", "dplyr", "factoextra", "FactoMineR", "tidyverse", "tibble", "reshape2", "gplots") #add new libraries here 
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load all libraries 
lapply(list.of.packages, FUN = function(X) {
  do.call("require", list(X)) 
})
source("../resources/biostats.R")
```

#### Read in dataframe with MACAU assocation results 

```{r}
macau.20190812 <- read.table(file="../analyses/macau/output/20190812-macau.assoc.txt", sep = "\t", stringsAsFactors = F)
head(macau.20190812)
colnames(macau.20190812) <- macau.20190812[1,]
macau.20190812 <- macau.20190812[-1, ] 
macau.20190812$pvalue <- as.numeric(macau.20190812$pvalue)
hist(macau.20190812$pvalue)
tail(macau.20190812)
```

#### Extract IDs for loci where MACAU indicates DML, where p<0.05 

```{r}
head(macau.20190812.05 <- subset(macau.20190812, pvalue<0.05))
```

#### Read in count data 

```{r}
counts.tot.destrand <- read.table(file = "../analyses/counts-total-destrand.txt", header = T)
counts.meth.destrand <- read.table(file = "../analyses/counts-meth-destrand.txt", header = T)
```

#### Extract raw count data for loci where p<0.05 in MACAU 

```{r}
# merge macau results with total counts 
head(macau.20190812.tot.counts <- merge(x=macau.20190812.05, y=subset(counts.tot.destrand, siteID %in% macau.20190812.05$id), by.x="id", by.y="siteID"))

# merge macau results with methylated counts 
head(macau.20190812.met.counts <- merge(x=macau.20190812.05, y=subset(counts.meth.destrand, siteID %in% macau.20190812.05$id), by.x="id", by.y="siteID"))

nrow(macau.20190812.05) - nrow(macau.20190812.met.counts) #3 row diff (due to duplicate contig ID's, weird)
```

#### Filter only loci with 10x or greater coverage for all samples 

```{r}
macau.20190812.tot.counts.10x <- macau.20190812.tot.counts %>% 
  filter_at(vars(starts_with("coverage")), all_vars(. > 9))

# Subset methylated count data / macau results for loci with 10x or greater coverage 
macau.20190812.met.counts.10x <- subset(macau.20190812.met.counts, id %in% macau.20190812.tot.counts.10x$id)
nrow(macau.20190812.met.counts.10x) - nrow(macau.20190812.tot.counts.10x) #2 extra rows in methyl.counts.10x df 

# Find duplicated contig entries in methyl.counts.10x df 
z <- macau.20190812.met.counts.10x[duplicated(macau.20190812.met.counts.10x$id),"id"]
macau.20190812.met.counts.10x[macau.20190812.met.counts.10x$id %in% z,] # Weird, duplicate entries have diff't counts

#Remove rows with the duplicate contigs 
macau.20190812.met.counts.10x <- macau.20190812.met.counts.10x[!(macau.20190812.met.counts.10x$id %in% z), ]
macau.20190812.tot.counts.10x <- macau.20190812.tot.counts.10x[!(macau.20190812.tot.counts.10x$id %in% z),]

# Confirm meth.10x and tot.10x df are same 
identical(macau.20190812.tot.counts.10x$id, macau.20190812.met.counts.10x$id) #yes 
```

#### Create % methylated dataframe from loci 10x coverage and MACAU p-value <0.05

```{r}
# duplicate macau methylated count dataframe to make % methylated df 
macau.20190812.perc.meth.10x <- macau.20190812.met.counts.10x

# divide macau methylated counts by total counts 
macau.20190812.perc.meth.10x[,15:32] <- macau.20190812.met.counts.10x[,15:32]/macau.20190812.tot.counts.10x[,15:32]
head(macau.20190812.perc.meth.10x[15:32])
```

### Visualize 

#### Heatmaps using raw counts  

---> mbd samples #1-9 = Hood Canal population  
---> mbd samples #10-18 = South Sound population  

```{r}
par(cex.main=.8)
heatmap(as.matrix(subset(macau.20190812.met.counts.10x, pvalue<0.01)[15:32]), main = "MACAU id'd counts, pvalue <0.01", xlab = "mbd sample", ylab = "loci")

heatmap(as.matrix(subset(macau.20190812.met.counts.10x, pvalue<0.001)[15:32]), main = "MACAU id'd counts, pvalue <0.001", xlab = "mbd sample", ylab = "loci")

heatmap(as.matrix(subset(macau.20190812.met.counts.10x, pvalue<0.0001)[15:32]), main = "MACAU id'd counts, pvalue <0.0001", xlab = "mbd sample", ylab = "loci")
```

#### PCA using raw counts 

```{r}
# Transpose methylated count df 
macau.20190812.met.counts.10x.t <- macau.20190812.met.counts.10x[c(1,15:32)] %>% remove_rownames %>% gather(column, value, -id) %>% spread(id, value) %>% column_to_rownames(var="column")

# Conduct PCA with variance-covariance matrix 
met.count.pca <- prcomp(macau.20190812.met.counts.10x.t, scale=F) #scale=F for variance-covariance matrix

# Assess PCA results 
summary(met.count.pca)
pca.eigenval(met.count.pca) #The Proporation of Variance = %variance 
fviz_screeplot(met.count.pca, addlabels = TRUE) #PC1 is sign., none other are  
fviz_contrib(met.count.pca, choice = "var", axes = 1, top=50) #check out loci that are high contributors
```

#### PC plot using PC1 and PC2   

```{r}
pops <- c("HC", rep("SS", times=9), rep("HC", times=8)) #create pop vector for color-coding 
fviz_pca_ind(met.count.pca, label="ind", col.ind=pops)+ ggtitle(label="PCA methylated counts, loci id'd by MACAU, PC 1x2")  #(could also use fviz_pca_biplot) 
```

#### Heatmaps using % methylated 

```{r}
par(cex.main=.8)
heatmap(as.matrix(subset(macau.20190812.perc.meth.10x, pvalue<0.01)[15:32]), main = "MACAU id'd % methylated, pvalue <0.01", xlab = "mbd sample", ylab = "loci")

heatmap(as.matrix(subset(macau.20190812.perc.meth.10x, pvalue<0.001)[15:32]), main = "MACAU id'd % methylated, pvalue <0.001", xlab = "mbd sample", ylab = "loci")

heatmap(as.matrix(subset(macau.20190812.perc.meth.10x, pvalue<0.0001)[15:32]), main = "MACAU id'd % methylated, pvalue <0.0001", xlab = "mbd sample", ylab = "loci")
```

#### PCA with % methylated df 

```{r}
# Transpose methylated count df 
macau.20190812.perc.meth.10x.t <- macau.20190812.perc.meth.10x[c(1,15:32)] %>% remove_rownames %>% gather(column, value, -id) %>% spread(id, value) %>% column_to_rownames(var="column")

met.perc.pca <- prcomp(macau.20190812.perc.meth.10x.t, scale=F) #scale=F for variance-covariance matrix
summary(met.perc.pca)
pca.eigenval(met.perc.pca) #The Proporation of Variance = %variance 
fviz_screeplot(met.perc.pca, addlabels = TRUE) #PC 1, 2 & 3 look sign.
fviz_contrib(met.perc.pca, choice = "var", axes = 1, top=50)

fviz_pca_biplot(met.perc.pca, label="ind", invisible = "var", col.ind=pops, axes = c(1,2), pointsize=3) + ggtitle(label="PCA % methylated, loci id'd by MACAU, PC 1x2")

fviz_pca_biplot(met.perc.pca, label="ind",  invisible = "var",col.ind=pops, axes = c(1,3), pointsize=3) + ggtitle(label="PCA % methylated, loci id'd by MACAU, PC 1x3")

fviz_pca_biplot(met.perc.pca, label="ind", invisible = "var", col.ind=pops, axes = c(2,3), pointsize=3) + ggtitle(label="PCA % methylated, loci id'd by MACAU, PC 2x3")  #dimension 2 = pop separation 
```